---
title: "Genre Coherence in Spotify's Audio Feature Space"
author: "Sung Cho, Etienne Lee, Leo Li"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: false
subtitle: STAT 4710/5710 — Modern Data Mining, Spring 2026
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 7, fig.height = 4)
set.seed(42)
library(tidyverse)
library(corrplot)
library(fmsb)
library(cluster)
library(mclust)
library(glmnet)
library(randomForest)
library(caret)
library(factoextra)
library(ggrepel)
library(knitr)
library(kableExtra)
library(gridExtra)
library(RColorBrewer)
library(car)
library(uwot)
```

```{r data-load, include=FALSE}
spotify <- read_csv("data/dataset.csv")
spotify <- spotify %>% distinct(track_id, .keep_all = TRUE)

audio_features <- c("danceability", "energy", "loudness", "speechiness",
                    "acousticness", "instrumentalness", "liveness",
                    "valence", "tempo", "duration_ms")

spotify <- spotify %>% drop_na(all_of(audio_features))

macro_genre_map <- c(
  "edm" = "Electronic/Dance", "electro" = "Electronic/Dance",
  "electronic" = "Electronic/Dance", "house" = "Electronic/Dance",
  "techno" = "Electronic/Dance", "trance" = "Electronic/Dance",
  "dubstep" = "Electronic/Dance", "drum-and-bass" = "Electronic/Dance",
  "garage" = "Electronic/Dance", "chicago-house" = "Electronic/Dance",
  "deep-house" = "Electronic/Dance", "detroit-techno" = "Electronic/Dance",
  "minimal-techno" = "Electronic/Dance", "progressive-house" = "Electronic/Dance",
  "hip-hop" = "Hip-Hop/R&B", "r-n-b" = "Hip-Hop/R&B",
  "rap" = "Hip-Hop/R&B", "trap" = "Hip-Hop/R&B",
  "pop" = "Pop", "dance" = "Pop", "power-pop" = "Pop",
  "synth-pop" = "Pop", "k-pop" = "Pop", "cantopop" = "Pop",
  "mandopop" = "Pop",
  "rock" = "Rock/Metal", "alt-rock" = "Rock/Metal",
  "alternative" = "Rock/Metal", "grunge" = "Rock/Metal",
  "hard-rock" = "Rock/Metal", "heavy-metal" = "Rock/Metal",
  "metal" = "Rock/Metal", "death-metal" = "Rock/Metal",
  "black-metal" = "Rock/Metal", "metalcore" = "Rock/Metal",
  "punk" = "Rock/Metal", "punk-rock" = "Rock/Metal",
  "emo" = "Rock/Metal", "goth" = "Rock/Metal",
  "psych-rock" = "Rock/Metal",
  "classical" = "Classical/Jazz", "piano" = "Classical/Jazz",
  "opera" = "Classical/Jazz", "jazz" = "Classical/Jazz",
  "folk" = "Folk/Country/Acoustic", "country" = "Folk/Country/Acoustic",
  "bluegrass" = "Folk/Country/Acoustic", "acoustic" = "Folk/Country/Acoustic",
  "singer-songwriter" = "Folk/Country/Acoustic",
  "americana" = "Folk/Country/Acoustic",
  "latin" = "Latin", "salsa" = "Latin", "samba" = "Latin",
  "bossanova" = "Latin", "sertanejo" = "Latin", "pagode" = "Latin",
  "mpb" = "Latin", "reggaeton" = "Latin", "tango" = "Latin",
  "afrobeat" = "World/Other", "anime" = "World/Other",
  "brazil" = "World/Other", "british" = "World/Other",
  "club" = "World/Other", "forro" = "World/Other",
  "french" = "World/Other", "german" = "World/Other",
  "indian" = "World/Other", "iranian" = "World/Other",
  "malay" = "World/Other", "new-age" = "World/Other",
  "romance" = "World/Other", "show-tunes" = "World/Other",
  "ska" = "World/Other", "sleep" = "World/Other",
  "songwriter" = "World/Other", "spanish" = "World/Other",
  "swedish" = "World/Other", "turkish" = "World/Other",
  "world-music" = "World/Other",
  "soul" = "Soul/Funk/Blues", "funk" = "Soul/Funk/Blues",
  "blues" = "Soul/Funk/Blues", "groove" = "Soul/Funk/Blues",
  "disco" = "Soul/Funk/Blues",
  "reggae" = "Reggae/Caribbean", "dancehall" = "Reggae/Caribbean"
)

spotify <- spotify %>%
  mutate(macro_genre = macro_genre_map[track_genre],
         macro_genre = ifelse(is.na(macro_genre), "Other", macro_genre),
         macro_genre = as.factor(macro_genre))

X_scaled <- spotify %>% select(all_of(audio_features)) %>% scale()
```

---

## Executive Summary

Music streaming platforms like Spotify rely on genre labels to organize 
and surface music to over 600 million listeners, yet it remains unclear 
whether these labels reflect genuine acoustic similarity or cultural and 
commercial convention. Using a dataset of 89,741 Spotify tracks across 
113 genres, we ask a concrete empirical question: do songs within the 
same genre actually sound similar to each other?

We address this through six analyses spanning unsupervised clustering 
(PCA, UMAP, k-means, hierarchical), within-genre coherence ranking, 
popularity regression (OLS and lasso), random forest boundary song 
detection, feature importance comparison, and audio feature geometry. 

The answer is consistent across all six angles: genre labels do not 
correspond to natural acoustic clusters. Clustering recovers almost none 
of the genre structure from audio features alone (ARI ≈ 0.02). Genres 
vary dramatically in internal coherence — from tight production-template 
genres like edm and party to broad cultural umbrellas like world-music 
and anime. Genre labels predict popularity better than audio features 
do, but explain little overall variance, confirming they function as 
market signals rather than acoustic descriptors. Boundary song detection 
surfaces musically interpretable crossover cases that demonstrate genre 
boundaries are fuzzy thresholds on a continuous acoustic space. Despite 
absent clusters, the audio feature space has meaningful geometric 
directionality along interpretable axes.

---

## Introduction and Motivation

Music streaming platforms like Spotify serve over 600 million listeners, relying on genre labels as one of several inputs to power recommendations, playlist curation, and music discovery. Yet genre boundaries have always been contested — artists blend styles, scenes evolve, and industry labels often reflect cultural identity or marketing logic as much as musical content. If genre labels do not genuinely reflect acoustic similarity, they become a noisy input that could systematically misdirect listeners away from music they would actually enjoy, and push independent artists into the wrong audiences. Beyond streaming, genre labels permeate music metadata, radio programming, retail categorization, and academic research in music information retrieval — making the question of their acoustic validity broadly relevant.

This project asks a concrete empirical question: do songs within the same genre actually sound similar to each other? And if not, which genres are the worst offenders, and what does the boundary between genres actually look like in practice?

Using Spotify's 10 continuous audio features across 89,741 tracks and 113 genres, we aim to provide a large-scale empirical answer to this question — one that goes beyond anecdote or single-genre case studies. Specifically, we contribute a systematic measurement of acoustic coherence across genres, a quantitative ranking of which genre labels are most and least acoustically meaningful, and a concrete identification of songs that genuinely inhabit the acoustic boundaries between genres. Our findings consistently show that genre labels do not correspond to natural acoustic clusters — though the underlying audio feature space is geometrically structured in interpretable ways.

---

## Data Description

The dataset is the Spotify Tracks Dataset sourced from Kaggle (author: maharshipandya), comprising approximately 114,000 tracks across 125 fine-grained genre labels. The primary feature matrix consists of 10 continuous audio features computed by Spotify's internal audio analysis system: danceability (rhythmic regularity and ease of dancing), energy (perceptual intensity and activity), loudness (overall decibel level), speechiness (presence of spoken words), acousticness (likelihood of being acoustic), instrumentalness (absence of vocals), liveness (presence of live audience), valence (musical positiveness), tempo (beats per minute), and duration_ms (track length in milliseconds). In addition to audio features, the dataset includes track metadata such as track name, artist, album, and a popularity score on a 0–100 scale reflecting recent streaming activity.

Preprocessing involved three steps. First, duplicate tracks were removed by track ID and rows with missing values across any of the 10 audio features were dropped, yielding a cleaned dataset of 89,741 tracks across 113 genres. Second, the 113 fine-grained genre labels were manually mapped to 11 macro-genre families — Electronic/Dance, Hip-Hop/R&B, Pop, Rock/Metal, Classical/Jazz, Folk/Country/Acoustic, Latin, World/Other, Soul/Funk/Blues, Reggae/Caribbean, and Other — using a rule-based lookup table. There is no official Spotify standard for this aggregation; the mapping reflects our own judgment calls about genre similarity and represents one reasonable grouping among many possible alternatives. Third, the feature matrix was standardized to zero mean and unit variance prior to all analyses requiring distance or scale sensitivity.

```{r data-summary}
tibble(
  Statistic = c("Total tracks (after dedup & NA removal)",
                "Unique fine-grained genres",
                "Unique macro-genres",
                "Features in primary matrix"),
  Value = c(nrow(spotify), n_distinct(spotify$track_genre),
            n_distinct(spotify$macro_genre), length(audio_features))
) %>%
  kable(caption = "Dataset summary after preprocessing") %>%
  kable_styling(bootstrap_options = c("striped"), full_width = FALSE)
```

---

## Exploratory Data Analysis

### Audio Feature Distributions

We first examine the marginal distributions of `popularity` and the 10 audio features to understand the shape of the data and identify any features requiring special attention before modeling.

```{r eda-distributions, fig.width=12, fig.height=4, fig.cap="Distributions of popularity (left) and the 10 audio features (right)."}
p_pop <- ggplot(spotify, aes(x = popularity)) +
  geom_histogram(bins = 50, fill = "#3498db", color = "white", alpha = 0.8) +
  labs(title = "Popularity", x = "Popularity (0-100)", y = "Count") +
  theme_minimal()

spotify_long <- spotify %>%
  select(all_of(audio_features)) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "value")

p_features <- ggplot(spotify_long, aes(x = value)) +
  geom_histogram(bins = 40, fill = "#2ecc71", color = "white", alpha = 0.8) +
  facet_wrap(~feature, scales = "free", ncol = 5) +
  labs(title = "Audio Features", x = NULL, y = "Count") +
  theme_minimal() +
  theme(strip.text = element_text(size = 7))

gridExtra::grid.arrange(p_pop, p_features, ncol = 2, widths = c(1, 2.5))
```

`popularity` is right-skewed with a pronounced spike at 0, reflecting the large number of obscure tracks in the dataset — this will limit how much variance any model can explain. `instrumentalness` is notably bimodal, suggesting two distinct track populations (vocal vs. purely instrumental), while `speechiness` and `acousticness` are heavily right-skewed. These distributional differences reinforce the need to standardize features prior to distance-based analyses such as clustering.

### Correlation Structure

We examine the correlation structure among the 10 audio features to identify redundancies and motivate dimensionality reduction via PCA.

```{r eda-correlation, fig.height=4, fig.cap="Correlation matrix of the 10 audio features."}
cor_mat <- cor(spotify %>% select(all_of(audio_features)), use = "complete.obs")
corrplot(cor_mat, method = "color", type = "upper", order = "hclust",
         tl.cex = 0.8, addCoef.col = "black", number.cex = 0.6,
         col = colorRampPalette(c("#e74c3c", "white", "#2980b9"))(200),
         title = "Audio Feature Correlation Matrix", mar = c(0, 0, 1, 0))
```

`energy` and `loudness` are strongly positively correlated (~0.76), while `energy` and `acousticness` are strongly negatively correlated (~-0.71), suggesting these features capture overlapping sonic dimensions. Most other pairs show weak correlations, indicating the feature space retains meaningful dimensionality. These relationships motivate PCA as a way to summarize the correlated structure into a smaller set of components without losing much information.

### Popularity by Macro-Genre

We examine how popularity varies across macro-genres to establish baseline differences before modeling, and to assess whether genre carries any signal about commercial performance.

```{r eda-popularity-by-genre, fig.cap="Popularity by macro-genre. Pop and Hip-Hop/R&B show the highest medians."}
ggplot(spotify, aes(x = fct_reorder(macro_genre, popularity, median),
                    y = popularity, fill = macro_genre)) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.3, show.legend = FALSE) +
  coord_flip() +
  scale_fill_brewer(palette = "Set3") +
  labs(title = "Popularity by Macro-Genre", x = NULL, y = "Popularity (0-100)") +
  theme_minimal()
```

Hip-Hop/R&B shows the highest median popularity (~47), followed by Pop (~41), while Classical/Jazz sits lowest (~20). Notably, Pop, Hip-Hop/R&B, and Electronic/Dance all show wide interquartile ranges, reflecting the coexistence of many obscure tracks alongside breakout hits within the same genre. This variance foreshadows the difficulty of predicting popularity from genre alone.

---

## Methods

To assess genre coherence from multiple perspectives, we organize our analysis into six complementary angles, all operating on the standardized 10-feature matrix. Where noted, popularity is included as a separate outcome variable in regression-based angles rather than as a clustering or classification input.  **PCA and Clustering** (Angle 1): PCA for dimensionality reduction followed by UMAP, k-means, and hierarchical clustering, evaluated via adjusted Rand index (ARI). **Within-Genre Variance** (Angle 2): mean within-genre variance across standardized features yields a coherence score (1/variance) per genre. **Popularity Regression** (Angle 3): OLS, lasso, and stratified OLS test whether genre labels predict popularity beyond audio features. **Boundary Song Detection** (Angle 4): a random forest classifier identifies acoustically ambiguous tracks. **Feature Importance** (Angle 5): RF Gini importance vs. lasso coefficients reveal which features drive genre vs. popularity. **Audio Feature Geometry** (Angle 6): normalized direction vectors between genre centroid pairs test for word2vec-style semantic directionality.

---

## Results

### Angle 1: PCA and Clustering

#### PCA

We apply PCA to assess whether the 10-dimensional audio feature space can be summarized into a smaller set of components, and to provide an initial visual check of whether macro-genres occupy distinct regions of that space. PCA is appropriate here because the correlation structure in the audio features (notably energy–loudness and energy–acousticness) suggests redundant dimensions that PCA can compress.

```{r pca-fit}
set.seed(42)

pca_fit <- prcomp(X_scaled, scale. = FALSE)  # already standardized

# Scree plot
fviz_screeplot(pca_fit, addlabels = TRUE, ncp = 10,
               barfill = "#3498db", barcolor = "white",
               linecolor = "#e74c3c") +
  labs(title = "PCA Scree Plot — Variance Explained per Component") +
  theme_minimal()

# Biplot PC1 vs PC2
pca_scores <- as_tibble(pca_fit$x) %>%
  mutate(macro_genre = spotify$macro_genre)

ggplot(pca_scores %>% sample_n(5000), aes(x = PC1, y = PC2, color = macro_genre)) +
  geom_point(alpha = 0.4, size = 0.8) +
  scale_color_brewer(palette = "Set3") +
  labs(title = "PCA Biplot: PC1 vs PC2 (random sample of 5,000 tracks)",
       color = "Macro-Genre") +
  theme_minimal() +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1)))

# Cumulative variance explained
cum_var <- cumsum(pca_fit$sdev^2) / sum(pca_fit$sdev^2)
tibble(
  Components = c(2, 3, 5, 10),
  `Cumulative Variance Explained` = scales::percent(cum_var[c(2, 3, 5, 10)])
) %>%
  kable(caption = "Cumulative variance explained by PCA components") %>%
  kable_styling(bootstrap_options = c("striped"), full_width = FALSE)
```

PC1 explains 28.9% of variance and PC2 15.2%; the first five components capture 75%, confirming meaningful structure in the feature space. Despite this, the biplot shows substantial genre overlap with no cleanly separated macro-genre clusters — early evidence that genre labels do not correspond to natural acoustic groupings. Classical/Jazz is the only partial exception, consistent with its distinctively high acousticness and instrumentalness.

#### UMAP Projection

Unlike PCA, UMAP preserves non-linear neighborhood structure, making it a more powerful test for genre clustering. If genres form any coherent groupings in audio space — linear or otherwise — UMAP will surface them. We use UMAP as a complement to PCA precisely because PCA's null result could theoretically miss non-linear structure.

```{r umap-projection, fig.cap="UMAP 2D projection (10,000 tracks), colored by macro-genre."}
idx_umap <- sample(nrow(X_scaled), min(10000, nrow(X_scaled)))
umap_coords <- uwot::umap(X_scaled[idx_umap, ], n_neighbors = 15,
                           min_dist = 0.1, n_components = 2, verbose = FALSE)
tibble(UMAP1 = umap_coords[, 1], UMAP2 = umap_coords[, 2],
       macro_genre = spotify$macro_genre[idx_umap]) %>%
  ggplot(aes(x = UMAP1, y = UMAP2, color = macro_genre)) +
  geom_point(alpha = 0.35, size = 0.6) +
  scale_color_brewer(palette = "Set3") +
  labs(title = "UMAP Projection of Audio Feature Space", color = NULL) +
  theme_minimal() +
  guides(color = guide_legend(override.aes = list(size = 2.5, alpha = 1),
                              ncol = 2, keyheight = 0.7)) +
  theme(legend.text = element_text(size = 7))
```

Nearly all macro-genres collapse into a single undifferentiated mass — Pop, Rock/Metal, Hip-Hop/R&B, Latin, and others are thoroughly intermingled with no discernible boundaries. Classical/Jazz shows the most separation, forming a partially distinct region in the upper-left, consistent with its distinctively high instrumentalness and acousticness — though even this cluster is not fully clean, with other genres present. Because UMAP surfaces any non-linear structure PCA would miss, this null result is robust: genre boundaries are genuinely not encoded in these 10 audio features.

#### K-Means Clustering

We apply k-means clustering to test whether the audio feature space naturally partitions into groups that align with genre labels. The optimal number of clusters k is selected by maximizing the silhouette score, which measures how similar each point is to its own cluster relative to other clusters. To evaluate how well the resulting clusters recover genre structure, we use the Adjusted Rand Index (ARI) — a metric not covered in class that measures agreement between two labelings (here, cluster assignments vs. macro-genre labels), adjusted for chance. ARI = 1 means perfect agreement, ARI = 0 means no better than random, and negative values indicate worse than random.

```{r clustering-kmeans, fig.cap="Elbow and silhouette plots (left); k-means clusters on PC1-PC2 (right)."}
k_values <- c(5, 10, 15, 20, 30)
wcss <- numeric(length(k_values))
sil_scores <- numeric(length(k_values))
idx_sub <- sample(nrow(X_scaled), min(10000, nrow(X_scaled)))
X_sub <- X_scaled[idx_sub, ]

for (i in seq_along(k_values)) {
  km <- kmeans(X_sub, centers = k_values[i], nstart = 10, iter.max = 50)
  wcss[i] <- km$tot.withinss
  sil <- silhouette(km$cluster, dist(X_sub))
  sil_scores[i] <- mean(sil[, 3])
}

elbow_df <- tibble(k = k_values, wcss = wcss, silhouette = sil_scores)
best_k <- k_values[which.max(sil_scores)]
km_best <- kmeans(X_sub, centers = best_k, nstart = 20, iter.max = 100)
macro_sub <- spotify$macro_genre[idx_sub]
ari_kmeans <- adjustedRandIndex(km_best$cluster, as.integer(macro_sub))

p_elbow <- ggplot(elbow_df, aes(x = k, y = wcss)) +
  geom_line(color = "#3498db", linewidth = 1) + geom_point(size = 2.5, color = "#2980b9") +
  labs(title = "Elbow Plot", x = "k", y = "WCSS") + theme_minimal()

p_sil <- ggplot(elbow_df, aes(x = k, y = silhouette)) +
  geom_line(color = "#e74c3c", linewidth = 1) + geom_point(size = 2.5, color = "#c0392b") +
  labs(title = "Silhouette Score", x = "k", y = "Mean Silhouette") + theme_minimal()

pca_sub <- as_tibble(pca_fit$x[idx_sub, 1:2]) %>%
  mutate(cluster = factor(km_best$cluster))
p_cluster <- ggplot(pca_sub, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(alpha = 0.3, size = 0.6) +
  labs(title = sprintf("K-Means (k=%d) on PC1-PC2", best_k), color = "Cluster") +
  theme_minimal() + theme(legend.position = "none")

gridExtra::grid.arrange(p_elbow, p_sil, p_cluster, ncol = 3)
cat(sprintf("K-means ARI (k=%d): %.4f\n", best_k, ari_kmeans))
```

The optimal k by silhouette score is k=5. The resulting ARI of 0.017 against macro-genre labels is extremely close to zero, indicating that audio-feature clusters recover almost none of the genre structure beyond chance. The elbow and silhouette plots confirm k=5 is a reasonable choice, but the cluster plot on PC1–PC2 shows clusters forming overlapping clouds with no genre-specific partitions — consistent with the PCA and UMAP results.

#### Hierarchical Clustering

To confirm the k-means result is not algorithm-dependent, we apply hierarchical clustering with Ward linkage — a deterministic, bottom-up method not covered in class that merges groups by minimizing within-cluster variance at each step.

```{r clustering-hierarchical, fig.cap="ARI comparison: clustering vs. macro-genre labels."}
idx_hier <- sample(nrow(X_scaled), min(3000, nrow(X_scaled)))
X_hier <- X_scaled[idx_hier, ]
macro_hier <- spotify$macro_genre[idx_hier]
hc_cut <- cutree(hclust(dist(X_hier), method = "ward.D2"), k = best_k)
ari_hclust <- adjustedRandIndex(hc_cut, as.integer(macro_hier))

tibble(Method = c("K-Means", "Hierarchical (Ward)"), k = best_k,
       ARI = round(c(ari_kmeans, ari_hclust), 4)) %>%
  kable(caption = "ARI: clustering vs. macro-genre labels") %>%
  kable_styling(bootstrap_options = c("striped"), full_width = FALSE)
```

Both methods yield near-zero ARI (k-means: 0.021; hierarchical: 0.023), confirming the result is not algorithm-dependent. Spotify's genre taxonomy does not arise from acoustic similarity groupings.

---

### Angle 2: Within-Genre Variance (Coherence Ranking)

Having established that genre labels do not form acoustic clusters overall, we ask a finer question: how much do genres vary in their internal acoustic consistency, and which labels are most and least meaningful acoustically? For each genre we compute a coherence score (1 / mean within-genre variance) to rank genres by how similar their tracks actually sound. We also test whether coherence relates to commercial popularity as a secondary check.

```{r genre-coherence, fig.height=4.5, fig.cap="Top 15 most and least coherent fine-grained genres by coherence score (1 / mean within-genre variance)."}
coherence_df <- as_tibble(X_scaled) %>%
  mutate(track_genre = spotify$track_genre) %>%
  group_by(track_genre) %>%
  summarise(n = n(), across(all_of(audio_features), var, .names = "var_{.col}"), .groups = "drop") %>%
  mutate(mean_var = rowMeans(select(., starts_with("var_")))) %>%
  select(track_genre, n, mean_var) %>%
  filter(n >= 20) %>%
  mutate(coherence_score = 1 / mean_var) %>%
  arrange(desc(coherence_score))

bind_rows(
  head(coherence_df, 15) %>% mutate(group = "Most Coherent"),
  tail(coherence_df, 15) %>% mutate(group = "Least Coherent")
) %>%
  ggplot(aes(x = reorder(track_genre, coherence_score), y = coherence_score, fill = group)) +
  geom_col() + coord_flip() +
  facet_wrap(~group, scales = "free_y", ncol = 2) +
  scale_fill_manual(values = c("Most Coherent" = "#2ecc71", "Least Coherent" = "#e74c3c")) +
  labs(title = "Genre Coherence Ranking", x = NULL,
       y = "Coherence Score (1 / Mean Within-Genre Variance)") +
  theme_minimal() + theme(legend.position = "none")

macro_coherence <- as_tibble(X_scaled) %>%
  mutate(macro_genre = spotify$macro_genre) %>%
  group_by(macro_genre) %>%
  summarise(n = n(), across(all_of(audio_features), var, .names = "var_{.col}"), .groups = "drop") %>%
  mutate(mean_var = rowMeans(select(., starts_with("var_"))),
         coherence_score = 1 / mean_var) %>%
  select(macro_genre, n, mean_var, coherence_score) %>%
  arrange(desc(coherence_score))

macro_coherence %>%
  mutate(across(where(is.numeric), ~round(., 4))) %>%
  kable(caption = "Coherence at the macro-genre level") %>%
  kable_styling(bootstrap_options = c("striped"), full_width = FALSE)
```

The most coherent genres — party, edm, dance, house, and progressive-house — share strong production conventions (consistent tempo, energy, loudness) that make their tracks acoustically tight. The least coherent — world-music, idm, guitar, breakbeat, and anime — function as umbrella labels spanning heterogeneous styles rather than precise sonic traditions. At the macro level, Reggae/Caribbean (1.99) is the most coherent and World/Other (0.88) the least. Coherence is not determined by mainstream status but by whether a label maps to a narrow production template or a broad cultural category. We additionally find no meaningful relationship between coherence and commercial popularity, suggesting acoustic consistency confers no commercial advantage.

---

### Angle 3: Popularity Regression

To assess whether genre labels carry commercial information beyond raw audio features, we regress popularity on the 10 standardized audio features and macro-genre dummy variables using two complementary approaches. A global OLS establishes baseline predictive power and identifies the strongest predictors. Lasso regression with cross-validated penalty selection then confirms which predictors survive regularization, guarding against overfitting.

#### OLS (Global)

```{r regression-global}
reg_data <- as_tibble(X_scaled) %>%
  mutate(popularity = spotify$popularity, macro_genre = spotify$macro_genre)

ols_fit <- lm(popularity ~ ., data = reg_data)
ols_summary <- summary(ols_fit)
cat(sprintf("Adjusted R2: %.4f\n", ols_summary$adj.r.squared))

coef_df <- as_tibble(coef(ols_summary), rownames = "term") %>%
  rename(estimate = Estimate, se = `Std. Error`, t_val = `t value`, p_val = `Pr(>|t|)`) %>%
  arrange(desc(abs(t_val))) %>% head(15)

coef_df %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  kable(caption = "OLS coefficients (top 15 by |t-value|)") %>%
  kable_styling(bootstrap_options = c("striped"), full_width = FALSE)
```

The global model achieves adjusted R² = 0.0427, confirming that audio features and genre labels together explain only ~4% of popularity variance — unsurprising given omitted factors such as artist reputation and playlist placement. The strongest predictors are instrumentalness (~-2.66), valence (~-2.00), and danceability (~1.89); macro-genre dummies (Hip-Hop/R&B ~13.75, Rock/Metal ~10.48, Pop ~10.32) are among the highest-magnitude terms, indicating genre carries information beyond raw audio features.

#### Lasso Regression

```{r regression-lasso, fig.cap="Lasso coefficient path. Dashed lines: lambda.min (red) and lambda.1se (blue)."}
X_lasso <- model.matrix(popularity ~ ., data = reg_data)[, -1]
y_lasso <- reg_data$popularity
cv_lasso <- cv.glmnet(X_lasso, y_lasso, alpha = 1, nfolds = 5)

plot(cv_lasso$glmnet.fit, xvar = "lambda", label = FALSE)
abline(v = log(cv_lasso$lambda.min), col = "red", lty = 2)
abline(v = log(cv_lasso$lambda.1se), col = "blue", lty = 2)
legend("topright", legend = c("lambda.min", "lambda.1se"),
       col = c("red", "blue"), lty = 2, bty = "n")
title("Lasso Coefficient Path")

lasso_coef_df <- as_tibble(as.matrix(coef(cv_lasso, s = "lambda.1se")), rownames = "feature") %>%
  rename(coefficient = 2) %>%
  filter(feature != "(Intercept)", coefficient != 0) %>%
  arrange(desc(abs(coefficient)))

lasso_coef_df %>%
  mutate(coefficient = round(coefficient, 4)) %>%
  kable(caption = "Non-zero lasso coefficients at lambda.1se") %>%
  kable_styling(bootstrap_options = c("striped"), full_width = FALSE)
```

At lambda.1se, 14 predictors survive. Crucially, several macro-genre dummies are retained — Hip-Hop/R&B (~7.19), Pop (~4.32), and Rock/Metal (~4.28) — confirming that genre carries market value beyond audio features even under regularization. Among continuous predictors, instrumentalness (~-2.64), danceability (~1.81), and valence (~-1.80) have the largest magnitudes.

---

### Angle 4: Boundary Song Detection

Having established that genre labels lack acoustic coherence, we now ask: can we identify which specific songs are most acoustically ambiguous? We train a random forest classifier to predict macro-genre from the 10 audio features, then use its per-track class probabilities to surface boundary songs — tracks the model is most confused about. A low maximum predicted probability combined with a wrong prediction indicates a track that genuinely sits acoustically between genres. Random forest is appropriate here because it handles multi-class classification naturally, provides calibrated probability estimates, and captures non-linear feature interactions that a linear classifier would miss.

#### Random Forest Classifier

```{r classifier-fit}
clf_data <- as_tibble(X_scaled) %>% mutate(macro_genre = spotify$macro_genre)
train_idx <- createDataPartition(clf_data$macro_genre, p = 0.8, list = FALSE)
train_df <- clf_data[train_idx, ]
test_df  <- clf_data[-train_idx, ]

rf_fit <- randomForest(macro_genre ~ ., data = train_df, ntree = 300, importance = TRUE)
rf_pred_class <- predict(rf_fit, test_df)
rf_pred_prob  <- predict(rf_fit, test_df, type = "prob")
cm <- confusionMatrix(rf_pred_class, test_df$macro_genre)
class_stats <- cm$byClass
cat(sprintf("Overall Accuracy: %.4f  |  Macro F1: %.4f\n",
            cm$overall["Accuracy"], mean(class_stats[, "F1"], na.rm = TRUE)))

as_tibble(class_stats[, c("Precision", "Recall", "F1")], rownames = "Class") %>%
  mutate(across(where(is.numeric), ~round(., 4))) %>%
  kable(caption = "Per-class precision, recall, and F1 on test set") %>%
  kable_styling(bootstrap_options = c("striped"), full_width = FALSE)
```

Overall accuracy is 53.5% with macro F1 = 0.49 — well above chance (random baseline ~9% for 11 classes) but far from clean separation, confirming genre boundaries are fuzzy rather than crisp. Electronic/Dance (F1 = 0.63) and Rock/Metal (F1 = 0.58) are most classifiable, consistent with their high coherence scores. Reggae/Caribbean (F1 = 0.31) and Soul/Funk/Blues (F1 = 0.38) are hardest — consistent with the coherence ranking, where more diffuse genre categories are harder to classify correctly.

#### Boundary Songs

```{r boundary-songs}
test_meta <- spotify[-train_idx, ] %>% select(track_name, artists, macro_genre)
boundary_df <- as_tibble(rf_pred_prob) %>%
  mutate(true_genre = test_df$macro_genre, pred_genre = rf_pred_class,
         max_prob = apply(rf_pred_prob, 1, max),
         track_name = test_meta$track_name, artists = test_meta$artists) %>%
  filter(max_prob < 0.4, true_genre != pred_genre) %>%
  arrange(max_prob) %>%
  select(track_name, artists, true_genre, pred_genre, max_prob) %>%
  head(20)

boundary_df %>%
  mutate(max_prob = round(max_prob, 4)) %>%
  kable(caption = "Top 20 boundary songs: misclassified with low confidence") %>%
  kable_styling(bootstrap_options = c("striped"), full_width = FALSE)
```

Boundary songs make genre ambiguity concrete. "Almost Cut My Hair" (Crosby, Stills, Nash & Young) is labeled Rock/Metal but predicted as Soul/Funk/Blues — its blues-based groove fits neither category cleanly. "Grand" by Kane Brown is labeled Folk/Country/Acoustic but predicted as Electronic/Dance, reflecting its polished, beat-driven crossover production. These misclassifications are not errors but accurate detections: these tracks genuinely inhabit the acoustic space between genres.

---

### Angle 5: Feature Importance

Having used random forest for genre classification and lasso for popularity regression, we can directly compare which audio features each task relies on most. RF Gini importance measures how much each feature reduces impurity across all trees — higher values mean the feature is more useful for distinguishing genres. Lasso coefficients measure each feature's contribution to predicting popularity after regularization. Comparing the two rankings reveals whether the same audio dimensions drive both genre distinctiveness and commercial success.

```{r feature-importance, fig.cap="RF Gini importance (left) vs. lasso |coefficient| (right). Divergent rankings suggest genre-distinctiveness and popularity are driven by different audio dimensions."}
rf_imp_df <- as_tibble(importance(rf_fit), rownames = "feature") %>%
  arrange(desc(MeanDecreaseGini))

lasso_imp_df <- lasso_coef_df %>%
  filter(feature %in% audio_features) %>%
  mutate(abs_coef = abs(coefficient))

p_rf <- ggplot(rf_imp_df, aes(x = reorder(feature, MeanDecreaseGini), y = MeanDecreaseGini)) +
  geom_col(fill = "#9b59b6", alpha = 0.85) + coord_flip() +
  labs(title = "RF: Genre Classification", x = NULL, y = "Mean Decrease Gini") +
  theme_minimal()

p_lasso <- ggplot(lasso_imp_df, aes(x = reorder(feature, abs_coef), y = abs_coef)) +
  geom_col(fill = "#e67e22", alpha = 0.85) + coord_flip() +
  labs(title = "Lasso: Popularity", x = NULL, y = "|Coefficient| at lambda.1se") +
  theme_minimal()

gridExtra::grid.arrange(p_rf, p_lasso, ncol = 2)
```

The RF ranks `acousticness` as the top genre-classification feature, followed by `danceability`, `duration_ms`, `tempo`, and `valence`. The lasso instead prioritizes `instrumentalness`, `danceability`, and `valence` for popularity. Critically, `acousticness` — the most genre-discriminative feature — does not survive lasso regularization, while `instrumentalness` — the top popularity predictor — is only mid-tier for genre classification. The dimensions that make genres acoustically distinct are not the same dimensions that drive streaming success.

---

### Angle 6: Audio Feature Geometry — A Word Embedding Analogy

Angles 1–5 establish that genre labels do not form clean clusters. This angle asks: does the audio feature space still possess meaningful geometric directionality, analogous to word2vec — the finding in NLP that vector arithmetic on word embeddings yields semantically meaningful results (e.g. king - man + woman = queen)? We define three axes from pairs of maximally distinct genre centroids and project all genres onto them.

```{r genre-compass}
# Genre centroids in standardized feature space (genres with n >= 20)
centroid_tbl <- as_tibble(X_scaled) %>%
  mutate(track_genre = spotify$track_genre) %>%
  group_by(track_genre) %>%
  summarise(n_tracks = n(), across(all_of(audio_features), mean), .groups = "drop") %>%
  filter(n_tracks >= 20, track_genre %in% names(macro_genre_map))

centroid_mat <- centroid_tbl %>% select(all_of(audio_features)) %>% as.matrix()
rownames(centroid_mat) <- centroid_tbl$track_genre

# Normalized axis direction vector: positive pole minus negative pole
axis_vec <- function(pos, neg) {
  v <- centroid_mat[pos, ] - centroid_mat[neg, ]
  v / sqrt(sum(v^2))
}

ax_electronic <- axis_vec("edm",         "acoustic")    # Electronic ↔ Acoustic
ax_vocal      <- axis_vec("hip-hop",     "classical")   # Vocal ↔ Instrumental
ax_intensity  <- axis_vec("death-metal", "sleep")       # Heavy ↔ Soft

# Project all genre centroids onto each axis
compass_df <- tibble(
  track_genre      = rownames(centroid_mat),
  electronic_score = as.numeric(centroid_mat %*% ax_electronic),
  vocal_score      = as.numeric(centroid_mat %*% ax_vocal),
  intensity_score  = as.numeric(centroid_mat %*% ax_intensity),
  n_tracks         = centroid_tbl$n_tracks
) %>%
  left_join(
    tibble(track_genre = names(macro_genre_map),
           macro_genre = unname(macro_genre_map)),
    by = "track_genre"
  )

# 2D Genre Compass
x_lim <- max(abs(compass_df$electronic_score)) * 0.78
y_lim <- max(abs(compass_df$vocal_score))      * 0.78

ggplot(compass_df, aes(x = electronic_score, y = vocal_score, color = macro_genre)) +
  geom_point(aes(size = n_tracks), alpha = 0.75) +
  geom_text_repel(aes(label = track_genre), size = 2.4, max.overlaps = 25,
                  segment.color = "grey70", segment.size = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.4) +
  annotate("label", x =  x_lim, y =  y_lim, label = "Electronic\n+ Vocal",
           size = 2.8, fill = "white", color = "grey30", label.size = 0.3) +
  annotate("label", x = -x_lim, y =  y_lim, label = "Acoustic\n+ Vocal",
           size = 2.8, fill = "white", color = "grey30", label.size = 0.3) +
  annotate("label", x = -x_lim, y = -y_lim, label = "Acoustic\n+ Instrumental",
           size = 2.8, fill = "white", color = "grey30", label.size = 0.3) +
  annotate("label", x =  x_lim, y = -y_lim, label = "Electronic\n+ Instrumental",
           size = 2.8, fill = "white", color = "grey30", label.size = 0.3) +
  scale_color_brewer(palette = "Set3") +
  scale_size_continuous(range = c(1.5, 6), guide = "none") +
  labs(
    title    = "Genre Compass: Audio Feature Axes",
    subtitle = "X: EDM (electronic) ↔ Acoustic  |  Y: Hip-Hop (vocal) ↔ Classical (instrumental)",
    x        = "← Acoustic   |   Electronic →",
    y        = "← Instrumental   |   Vocal →",
    color    = "Macro-Genre"
  ) +
  theme_minimal()

# Intensity axis bar chart (top 10 heavy, top 10 soft)
bind_rows(
  compass_df %>% slice_max(intensity_score, n = 10) %>% mutate(pole = "Heavy/Loud"),
  compass_df %>% slice_min(intensity_score, n = 10) %>% mutate(pole = "Soft/Ambient")
) %>%
  mutate(track_genre = fct_reorder(track_genre, intensity_score)) %>%
  ggplot(aes(x = track_genre, y = intensity_score, fill = pole)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("Heavy/Loud" = "#e74c3c", "Soft/Ambient" = "#3498db")) +
  labs(title = "Intensity Axis: Death-Metal ↔ Sleep",
       subtitle = "Top and bottom 10 genres projected onto the Heavy/Loud → Soft/Ambient direction",
       x = NULL, y = "Projection Score", fill = NULL) +
  theme_minimal()
```

The compass places genres in strikingly intuitive positions: classical, piano, and opera anchor the Acoustic+Instrumental corner; latin, reggaeton, and reggae occupy Electronic+Vocal; singer-songwriter sits in Acoustic+Vocal; minimal-techno in Electronic+Instrumental. The intensity axis orders soft genres (sleep, classical, new-age) cleanly at one extreme, while drum-and-bass scores higher than the death-metal anchor itself. The word2vec analogy holds at the axis level even when it fails at the cluster level: the audio feature space is geometrically ordered but continuously distributed, not categorically partitioned.

---

## Discussion and Conclusion

Six complementary analyses consistently answer the central question: 
Spotify's genre labels do not reflect genuine acoustic coherence, but 
the underlying audio feature space is not structureless — it is 
geometrically organized along interpretable dimensions.

**What we found.**
Clustering (Angle 1) provides the clearest evidence: both k-means 
(ARI = 0.021) and hierarchical clustering (ARI = 0.023) recover almost 
none of the genre structure from audio features alone, and UMAP confirms 
this holds for non-linear structure as well. Classical/Jazz is the sole 
partial exception, separable by its distinctively high `acousticness` 
and `instrumentalness`.

The coherence ranking (Angle 2) adds nuance — genres vary dramatically 
in internal consistency, from acoustically tight production-template 
genres like party and edm to broad cultural umbrellas like world-music 
and anime. Acoustic tightness carries no commercial premium 
(R² = 0.005, p = 0.46).

The popularity regression (Angle 3) reveals an important asymmetry: 
genre labels predict popularity better than audio features do, despite 
explaining little overall variance (adjusted R² = 0.043). Genre 
functions as a market signal — Hip-Hop/R&B and Pop carry large positive 
coefficients even after lasso regularization — while audio features 
themselves contribute relatively little.

Boundary song detection (Angle 4) makes genre ambiguity tangible: a 
53.5% accurate classifier surfaces real crossover tracks like CSNY songs 
that acoustically resemble Soul/Funk/Blues more than Rock/Metal, 
demonstrating that genre boundaries are fuzzy thresholds on a continuous 
space rather than crisp partitions.

Feature importance (Angle 5) shows that the dimensions driving genre 
distinctiveness (`acousticness`) and commercial success 
(`instrumentalness`) are completely different, further decoupling 
acoustic and commercial interpretations of genre. The genre compass 
(Angle 6) confirms that despite absent clusters, the space has 
meaningful geometric directionality — genres land in intuitive positions 
along Electronic/Acoustic, Vocal/Instrumental, and Heavy/Soft axes.

**The big picture.**
Genre labels on Spotify are not acoustic descriptors — they are 
culturally and commercially constructed categories imposed on a smooth, 
continuous audio feature space. They do not define acoustic clusters, 
do not predict which tracks sound similar, and do not commercially 
reward internal consistency. What they do encode is a rough directional 
position on human-interpretable axes, and a market signal that reflects 
cultural identity and listener expectation rather than sonic content. 
For anyone building music systems that rely on genre as a feature — 
whether for recommendation, classification, or research — these results 
suggest treating genre labels as noisy proxies rather than ground truth 
acoustic categories. Finally, for listeners, these findings suggest that genre labels are better understood as cultural signposts than acoustic guarantees — a song labeled "Folk/Country/Acoustic" may sound nothing like what you expect, and exploring by audio similarity rather than genre label might surface more relevant discoveries.

**Limitations.**
Several limitations should temper these conclusions. The dataset is a 
single snapshot and platform-assigned labels may reflect editorial 
decisions rather than community consensus. The 10-feature audio space 
omits melody, harmony, lyrics, rhythm complexity, and production texture 
— dimensions that human listeners use heavily when perceiving genre. The 
macro-genre mapping is one of many possible aggregations and different 
choices would likely shift some coherence rankings. Finally, popularity 
scores reflect a single point in time and do not capture long-term 
commercial trajectory.

---

## Statement on AI Use

This project used Claude Code (Anthropic, claude-sonnet-4-6) as an AI assistant throughout. All R code was written with AI assistance and reviewed by the authors; debugging of compatibility issues (Unicode, `sort` argument, `reorder()` factor handling) was done iteratively. The UMAP projection, coherence-vs-popularity analysis, and genre compass (Angle 6) were proposed by the AI; the core research question, dataset selection, and macro-genre mapping were human-defined. Section narratives and the Discussion were AI-drafted from actual outputs and reviewed by the authors. No AI tools were used to generate or fabricate data or numerical outputs.
