---
title: "Genre Coherence in Spotify's Audio Feature Space"
subtitle: "STAT 4710/5710 — Modern Data Mining, Spring 2026"
author: "Sung Cho, Etienne Lee, Leo Li"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
# Set working directory to the folder containing dataset.csv before knitting
# dataset source: https://www.kaggle.com/datasets/maharshipandya/-spotify-tracks-dataset

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5)

set.seed(42)

# install.packages(c(
#   "tidyverse", "corrplot", "fmsb", "cluster", "mclust",
#   "glmnet", "randomForest", "caret", "factoextra",
#   "ggrepel", "knitr", "kableExtra", "gridExtra",
#   "RColorBrewer", "car", "uwot"
# ))

library(tidyverse)
library(corrplot)
library(fmsb)
library(cluster)
library(mclust)
library(glmnet)
library(randomForest)
library(caret)
library(factoextra)
library(ggrepel)
library(knitr)
library(kableExtra)
library(gridExtra)
library(RColorBrewer)
library(car)
library(uwot)
```

```{r data-load, include=FALSE}
spotify <- read_csv("dataset.csv")

# Remove duplicate track_id entries — keep first occurrence
spotify <- spotify %>%
  distinct(track_id, .keep_all = TRUE)

# Define the 10 audio features used throughout
audio_features <- c("danceability", "energy", "loudness", "speechiness",
                    "acousticness", "instrumentalness", "liveness",
                    "valence", "tempo", "duration_ms")

# Drop rows with any NA in the 10 audio features
spotify <- spotify %>%
  drop_na(all_of(audio_features))

# Macro-genre mapping
macro_genre_map <- c(
  # Electronic / Dance
  "edm" = "Electronic/Dance", "electro" = "Electronic/Dance",
  "electronic" = "Electronic/Dance", "house" = "Electronic/Dance",
  "techno" = "Electronic/Dance", "trance" = "Electronic/Dance",
  "dubstep" = "Electronic/Dance", "drum-and-bass" = "Electronic/Dance",
  "garage" = "Electronic/Dance", "chicago-house" = "Electronic/Dance",
  "deep-house" = "Electronic/Dance", "detroit-techno" = "Electronic/Dance",
  "minimal-techno" = "Electronic/Dance", "progressive-house" = "Electronic/Dance",
  # Hip-Hop / R&B
  "hip-hop" = "Hip-Hop/R&B", "r-n-b" = "Hip-Hop/R&B",
  "rap" = "Hip-Hop/R&B", "trap" = "Hip-Hop/R&B",
  # Pop
  "pop" = "Pop", "dance" = "Pop", "power-pop" = "Pop",
  "synth-pop" = "Pop", "k-pop" = "Pop", "cantopop" = "Pop",
  "mandopop" = "Pop",
  # Rock / Metal
  "rock" = "Rock/Metal", "alt-rock" = "Rock/Metal",
  "alternative" = "Rock/Metal", "grunge" = "Rock/Metal",
  "hard-rock" = "Rock/Metal", "heavy-metal" = "Rock/Metal",
  "metal" = "Rock/Metal", "death-metal" = "Rock/Metal",
  "black-metal" = "Rock/Metal", "metalcore" = "Rock/Metal",
  "punk" = "Rock/Metal", "punk-rock" = "Rock/Metal",
  "emo" = "Rock/Metal", "goth" = "Rock/Metal",
  "psych-rock" = "Rock/Metal",
  # Classical / Jazz
  "classical" = "Classical/Jazz", "piano" = "Classical/Jazz",
  "opera" = "Classical/Jazz", "jazz" = "Classical/Jazz",
  # Folk / Country / Acoustic
  "folk" = "Folk/Country/Acoustic", "country" = "Folk/Country/Acoustic",
  "bluegrass" = "Folk/Country/Acoustic", "acoustic" = "Folk/Country/Acoustic",
  "singer-songwriter" = "Folk/Country/Acoustic",
  "americana" = "Folk/Country/Acoustic",
  # Latin
  "latin" = "Latin", "salsa" = "Latin", "samba" = "Latin",
  "bossanova" = "Latin", "sertanejo" = "Latin", "pagode" = "Latin",
  "mpb" = "Latin", "reggaeton" = "Latin", "tango" = "Latin",
  # World / Other
  "afrobeat" = "World/Other", "anime" = "World/Other",
  "brazil" = "World/Other", "british" = "World/Other",
  "club" = "World/Other", "forro" = "World/Other",
  "french" = "World/Other", "german" = "World/Other",
  "indian" = "World/Other", "iranian" = "World/Other",
  "malay" = "World/Other", "new-age" = "World/Other",
  "romance" = "World/Other", "show-tunes" = "World/Other",
  "ska" = "World/Other", "sleep" = "World/Other",
  "songwriter" = "World/Other", "spanish" = "World/Other",
  "swedish" = "World/Other", "turkish" = "World/Other",
  "world-music" = "World/Other",
  # Soul / Funk / Blues
  "soul" = "Soul/Funk/Blues", "funk" = "Soul/Funk/Blues",
  "blues" = "Soul/Funk/Blues", "groove" = "Soul/Funk/Blues",
  "disco" = "Soul/Funk/Blues",
  # Reggae / Caribbean
  "reggae" = "Reggae/Caribbean", "dancehall" = "Reggae/Caribbean"
)

spotify <- spotify %>%
  mutate(macro_genre = macro_genre_map[track_genre],
         macro_genre = ifelse(is.na(macro_genre), "Other", macro_genre),
         macro_genre = as.factor(macro_genre))

# Standardized feature matrix
X_scaled <- spotify %>%
  select(all_of(audio_features)) %>%
  scale()
```

---

## Executive Summary

Music streaming platforms assign genre labels to tracks as a primary means of organization and discovery, yet it remains unclear whether these labels reflect genuine musical coherence or function primarily as marketing categories. This project investigates the relationship between Spotify's genre taxonomy and the underlying audio feature space using a dataset of 114,000 tracks spanning 125 genres. We apply principal component analysis to reduce the high-dimensional audio feature space and use k-means and hierarchical clustering to assess whether unsupervised groupings recover genre labels, quantified by adjusted Rand index and silhouette score. We further compute within-genre variance across all 125 genres to produce a genre coherence ranking, train a multi-class classifier to identify boundary songs that are musically proximate to genres other than their assigned label, and fit lasso regression models — both globally and stratified by macro-genre — to assess whether genre label predicts track popularity beyond what audio features alone explain. Together, these analyses ask whether genre is a meaningful descriptor of musical identity or an artifact of industry convention, with implications for how streaming platforms structure discovery and recommendation.

---

## Introduction and Motivation

Genre is one of the most fundamental organizing principles in music. Streaming platforms like Spotify use genre labels to drive recommendation algorithms, playlist curation, and discovery features. Yet the boundary between genres has always been porous: artists blend styles, listeners resist categorization, and industry labels often reflect marketing logic rather than musical content. This project uses Spotify's audio feature API — a set of continuous measurements derived from signal processing — to ask a concrete question: do songs within the same genre actually sound similar to each other?

We approach this question from six angles: unsupervised clustering (does the audio feature space cluster into genre-like groups?), within-genre variance (which genres are musically tight vs. diffuse?), popularity regression (does genre predict popularity beyond what audio features explain?), boundary song detection (which songs are misclassified by a genre classifier, and what does that reveal?), feature importance (which audio dimensions drive genre separation vs. popularity?), and audio feature geometry (does the space have interpretable directional structure analogous to word embeddings?). Together these form a comprehensive audit of Spotify's genre taxonomy.

---

## Data Description

The dataset is the [Spotify Tracks Dataset](https://www.kaggle.com/datasets/maharshipandya/-spotify-tracks-dataset) from Kaggle (author: maharshipandya), comprising approximately 114,000 tracks across 125 fine-grained genres. Each row corresponds to a unique track with 20 columns: metadata (track ID, artist, album, name), one outcome variable (`popularity`, scored 0–100 based on recent play counts), three discrete audio features (`key`, `mode`, `time_signature`), and ten continuous audio features that form our primary feature matrix.

The ten continuous features used throughout this analysis are: `danceability`, `energy`, `loudness`, `speechiness`, `acousticness`, `instrumentalness`, `liveness`, `valence`, `tempo`, and `duration_ms`. The discrete features (`key`, `mode`, `time_signature`) are excluded from the primary feature matrix but are explored in EDA. After removing duplicate track IDs and rows with missing values in the audio features, the cleaned dataset contains `r nrow(spotify)` tracks across `r n_distinct(spotify$track_genre)` genres.

We map the 125 fine-grained genres to `r n_distinct(spotify$macro_genre)` macro-genre families (e.g., Pop, Rock/Metal, Electronic/Dance) to enable higher-level comparisons and visualizations. Any genres not covered by the mapping are labeled "Other."

```{r data-summary}
tibble(
  Statistic = c("Total tracks (after dedup & NA removal)",
                "Unique fine-grained genres",
                "Unique macro-genres",
                "Features in primary matrix"),
  Value = c(nrow(spotify),
            n_distinct(spotify$track_genre),
            n_distinct(spotify$macro_genre),
            length(audio_features))
) %>%
  kable(caption = "Dataset summary after preprocessing") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

## Exploratory Data Analysis

We begin with univariate and bivariate summaries of the audio features and genre distribution to understand the data structure before modeling.

### Distributions of Audio Features and Popularity

```{r eda-distributions, fig.cap="Distributions of popularity and the 10 audio features. Note the spike at popularity = 0 and the strong bimodality of instrumentalness."}
# Popularity histogram
p_pop <- ggplot(spotify, aes(x = popularity)) +
  geom_histogram(bins = 50, fill = "#3498db", color = "white", alpha = 0.8) +
  labs(title = "Distribution of Popularity", x = "Popularity (0–100)", y = "Count") +
  theme_minimal()

# Audio feature density plots
spotify_long <- spotify %>%
  select(all_of(audio_features)) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "value")

p_features <- ggplot(spotify_long, aes(x = value)) +
  geom_histogram(bins = 40, fill = "#2ecc71", color = "white", alpha = 0.8) +
  facet_wrap(~feature, scales = "free", ncol = 5) +
  labs(title = "Distributions of Audio Features", x = NULL, y = "Count") +
  theme_minimal() +
  theme(strip.text = element_text(size = 9))

print(p_pop)
print(p_features)
```

The `popularity` distribution is right-skewed with a pronounced spike at 0, reflecting the large number of obscure tracks. `instrumentalness` is highly bimodal: most tracks are either strongly vocal or strongly instrumental with few intermediate values. `speechiness` and `acousticness` are also right-skewed.

### Correlation Structure

```{r eda-correlation, fig.cap="Correlation matrix of the 10 audio features. Strong positive correlation between energy and loudness; strong negative correlation between energy and acousticness."}
cor_mat <- cor(spotify %>% select(all_of(audio_features)), use = "complete.obs")

corrplot(cor_mat,
         method = "color",
         type = "upper",
         order = "hclust",
         tl.cex = 0.85,
         addCoef.col = "black",
         number.cex = 0.65,
         col = colorRampPalette(c("#e74c3c", "white", "#2980b9"))(200),
         title = "Audio Feature Correlation Matrix",
         mar = c(0, 0, 1, 0))
```

As expected, `energy` and `loudness` are strongly positively correlated (~0.76), while `energy` and `acousticness` are strongly negatively correlated (~−0.71). `valence` and `danceability` show moderate positive correlation (~0.46). These relationships motivate dimensionality reduction via PCA.

### Genre Counts by Macro-Genre

```{r eda-genre-counts, fig.cap="Track counts by macro-genre. Some macro-genre families are substantially larger than others, reflecting the original dataset composition."}
spotify %>%
  count(macro_genre) %>%
  arrange(desc(n)) %>%
  ggplot(aes(x = reorder(macro_genre, n), y = n, fill = macro_genre)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_brewer(palette = "Set3") +
  labs(title = "Track Counts by Macro-Genre",
       x = "Macro-Genre", y = "Number of Tracks") +
  theme_minimal()
```

### Audio Fingerprint by Macro-Genre (Radar Chart)

```{r eda-radar, fig.cap="Radar charts showing the mean standardized audio profile ('fingerprint') for each macro-genre family.", fig.height=10}
# Compute mean standardized features per macro-genre
radar_data <- as_tibble(X_scaled) %>%
  mutate(macro_genre = spotify$macro_genre) %>%
  group_by(macro_genre) %>%
  summarise(across(everything(), mean), .groups = "drop")

genres_list <- levels(spotify$macro_genre)

# Radar chart requires max/min rows prepended
radar_mat <- radar_data %>%
  select(-macro_genre) %>%
  as.data.frame()
rownames(radar_mat) <- radar_data$macro_genre

max_vals <- apply(radar_mat, 2, max) + 0.1
min_vals <- apply(radar_mat, 2, min) - 0.1
radar_plot_data <- rbind(max_vals, min_vals, radar_mat)

par(mfrow = c(3, 4), mar = c(1, 1, 2, 1))
colors_radar <- RColorBrewer::brewer.pal(min(nrow(radar_mat), 11), "Set3")
for (i in seq_len(nrow(radar_mat))) {
  fmsb::radarchart(
    radar_plot_data[c(1, 2, i + 2), ],
    axistype = 1,
    pcol = colors_radar[(i - 1) %% length(colors_radar) + 1],
    pfcol = adjustcolor(colors_radar[(i - 1) %% length(colors_radar) + 1], alpha.f = 0.3),
    plwd = 2,
    cglcol = "grey80",
    cglty = 1,
    axislabcol = "grey50",
    vlcex = 0.65,
    title = rownames(radar_mat)[i]
  )
}
par(mfrow = c(1, 1))
```

### Popularity by Macro-Genre

```{r eda-popularity-by-genre, fig.cap="Distribution of popularity scores by macro-genre. Pop and Hip-Hop/R&B tend to have higher median popularity; Classical/Jazz and Electronic/Dance show wider variance."}
ggplot(spotify, aes(x = fct_reorder(macro_genre, popularity, median),
                    y = popularity, fill = macro_genre)) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.3, show.legend = FALSE) +
  coord_flip() +
  scale_fill_brewer(palette = "Set3") +
  labs(title = "Popularity by Macro-Genre",
       x = "Macro-Genre", y = "Popularity (0–100)") +
  theme_minimal()
```

[PLACEHOLDER: interpret which genres have higher median popularity vs. wider variance and connect to known streaming trends]

---

## Methods

This section describes the five analytical angles applied to the dataset. All methods operate on the standardized 10-feature matrix unless otherwise noted.

### PCA and Clustering

We use PCA to project the 10-dimensional audio feature space into a lower-dimensional space for visualization and as input to clustering. K-means and hierarchical clustering are then applied to assess whether unsupervised groupings align with genre labels.

### Within-Genre Variance

For each of the 125 fine-grained genres, we compute the mean within-genre variance across all standardized features, yielding a coherence score. Genres with low variance (high coherence) have tracks that sound similar to each other; diffuse genres have high internal variance.

### Popularity Regression

We model `popularity` as a function of audio features and macro-genre dummies using OLS, lasso (global), and stratified OLS per macro-genre. This tests whether genre labels carry market value beyond what audio features explain.

### Boundary Song Detection

A random forest classifier is trained to predict macro-genre from audio features. Tracks where the true genre receives low predicted probability — while another genre receives the highest probability — are termed "boundary songs" and reveal where genre assignments are musically ambiguous.

### Feature Importance

We synthesize classifier variable importance (random forest) with lasso coefficient magnitudes to ask which audio dimensions drive genre separation vs. popularity prediction.

### Audio Feature Geometry

Inspired by word2vec-style vector arithmetic, we define interpretable axes in the 10-dimensional audio feature space by computing normalized direction vectors between pairs of maximally distinct genre centroids. Projecting all genre centroids onto these axes tests whether the space has semantic directionality — genres sorting sensibly along acoustic, vocal, and intensity dimensions — even in the absence of clean cluster separation.

---

## Results

### Angle 1: PCA and Clustering

#### PCA

```{r pca-fit, fig.cap="Left: Scree plot showing variance explained by each principal component. Right: PCA biplot of PC1 vs. PC2 colored by macro-genre."}
set.seed(42)

pca_fit <- prcomp(X_scaled, scale. = FALSE)  # already standardized

# Scree plot
fviz_screeplot(pca_fit, addlabels = TRUE, ncp = 10,
               barfill = "#3498db", barcolor = "white",
               linecolor = "#e74c3c") +
  labs(title = "PCA Scree Plot — Variance Explained per Component") +
  theme_minimal()

# Biplot PC1 vs PC2
pca_scores <- as_tibble(pca_fit$x) %>%
  mutate(macro_genre = spotify$macro_genre)

ggplot(pca_scores %>% sample_n(5000), aes(x = PC1, y = PC2, color = macro_genre)) +
  geom_point(alpha = 0.4, size = 0.8) +
  scale_color_brewer(palette = "Set3") +
  labs(title = "PCA Biplot: PC1 vs PC2 (random sample of 5,000 tracks)",
       color = "Macro-Genre") +
  theme_minimal() +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1)))

# Cumulative variance explained
cum_var <- cumsum(pca_fit$sdev^2) / sum(pca_fit$sdev^2)
tibble(
  Components = c(2, 3, 5, 10),
  `Cumulative Variance Explained` = scales::percent(cum_var[c(2, 3, 5, 10)])
) %>%
  kable(caption = "Cumulative variance explained by PCA components") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

[PLACEHOLDER: interpret how much variance is captured by the first 2–5 PCs and whether macro-genres separate visually in the biplot]

#### UMAP Projection

Unlike PCA, which finds the linear directions of maximum variance, UMAP (Uniform Manifold Approximation and Projection) preserves local neighborhood structure in a non-linear 2D embedding. It often reveals cluster separation that PCA cannot show.

```{r umap-projection, fig.cap="UMAP 2D projection of the standardized audio feature space (10,000 tracks), colored by macro-genre. Compared to the PCA biplot, UMAP better preserves local neighborhood structure and can reveal non-linear separation between genre families."}
set.seed(42)

idx_umap <- sample(nrow(X_scaled), min(10000, nrow(X_scaled)))
X_umap_in <- X_scaled[idx_umap, ]
macro_umap <- spotify$macro_genre[idx_umap]

umap_coords <- uwot::umap(X_umap_in, n_neighbors = 15, min_dist = 0.1,
                           n_components = 2, verbose = FALSE)

umap_df <- tibble(
  UMAP1 = umap_coords[, 1],
  UMAP2 = umap_coords[, 2],
  macro_genre = macro_umap
)

ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = macro_genre)) +
  geom_point(alpha = 0.35, size = 0.7) +
  scale_color_brewer(palette = "Set3") +
  labs(title = "UMAP Projection of Audio Feature Space (10,000 tracks)",
       color = "Macro-Genre") +
  theme_minimal() +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1)))
```

The UMAP projection reveals a strikingly clear result: nearly all macro-genres collapse into a single large, undifferentiated mass. Pop, Rock/Metal, Hip-Hop/R&B, Latin, Folk/Country/Acoustic, Soul/Funk/Blues, and World/Other are thoroughly intermingled throughout the central blob, with no discernible boundaries between them even under UMAP's non-linear lens.

The one meaningful exception is Classical/Jazz, which forms a partially separated cluster in the upper-right region of the embedding. This aligns with its known audio fingerprint — distinctively high instrumentalness and acousticness, low energy and danceability — making it the only macro-genre that is genuinely acoustically coherent and spatially separable. A small isolated cluster also appears at the far bottom-left, consisting of tracks with unusual feature combinations far from the genre mainstream.

Crucially, this null result is more informative than it might appear. UMAP is the most powerful tool available for revealing hidden cluster structure — unlike PCA, it is not constrained to linear projections and will surface any non-linear geometry in the data. The fact that genres remain inseparable even under UMAP closes the strongest counterargument to the low ARI finding from clustering: there is no non-linear structure that PCA was simply missing. Genre boundaries are genuinely not encoded in these 10 audio features, with the single exception of Classical/Jazz.

#### K-Means Clustering

```{r clustering-kmeans, fig.cap="Elbow plot (WCSS) and silhouette scores for k-means across k = 5 to 30. The selected k balances the elbow heuristic and peak silhouette."}
set.seed(42)

k_values <- c(5, 10, 15, 20, 30)
wcss <- numeric(length(k_values))
sil_scores <- numeric(length(k_values))

# Subsample for speed (hierarchical will use same subsample)
set.seed(42)
idx_sub <- sample(nrow(X_scaled), min(10000, nrow(X_scaled)))
X_sub <- X_scaled[idx_sub, ]

for (i in seq_along(k_values)) {
  km <- kmeans(X_sub, centers = k_values[i], nstart = 10, iter.max = 50)
  wcss[i] <- km$tot.withinss
  sil <- silhouette(km$cluster, dist(X_sub))
  sil_scores[i] <- mean(sil[, 3])
}

elbow_df <- tibble(k = k_values, wcss = wcss, silhouette = sil_scores)

p_elbow <- ggplot(elbow_df, aes(x = k, y = wcss)) +
  geom_line(color = "#3498db", linewidth = 1) +
  geom_point(size = 3, color = "#2980b9") +
  labs(title = "Elbow Plot: WCSS vs. k", x = "k", y = "Within-Cluster SS") +
  theme_minimal()

p_sil <- ggplot(elbow_df, aes(x = k, y = silhouette)) +
  geom_line(color = "#e74c3c", linewidth = 1) +
  geom_point(size = 3, color = "#c0392b") +
  labs(title = "Silhouette Score vs. k", x = "k", y = "Mean Silhouette") +
  theme_minimal()

gridExtra::grid.arrange(p_elbow, p_sil, ncol = 2)

# Select best k
best_k <- k_values[which.max(sil_scores)]
cat("Selected k:", best_k, "\n")

# Final k-means at best k
set.seed(42)
km_best <- kmeans(X_sub, centers = best_k, nstart = 20, iter.max = 100)

# ARI against macro_genre
macro_sub <- spotify$macro_genre[idx_sub]
ari_kmeans <- adjustedRandIndex(km_best$cluster, as.integer(macro_sub))
cat(sprintf("K-means ARI (k=%d) vs. macro_genre: %.4f\n", best_k, ari_kmeans))

# Visualize on PC1-PC2
pca_sub <- as_tibble(pca_fit$x[idx_sub, 1:2]) %>%
  mutate(cluster = factor(km_best$cluster),
         macro_genre = macro_sub)

ggplot(pca_sub, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(alpha = 0.3, size = 0.7) +
  labs(title = sprintf("K-Means Clusters (k=%d) on PC1-PC2", best_k),
       color = "Cluster") +
  theme_minimal()
```

[PLACEHOLDER: interpret ARI value — note that ARI near 0 indicates near-random alignment, near 1 = perfect recovery. Low ARI is the expected and central finding.]

#### Hierarchical Clustering

```{r clustering-hierarchical, fig.cap="Hierarchical clustering (Ward's method) cut at the selected k. ARI compared against macro-genre labels and versus k-means."}
set.seed(42)

# Use smaller subsample for hierarchical (distance matrix is O(n^2))
idx_hier <- sample(nrow(X_scaled), min(3000, nrow(X_scaled)))
X_hier <- X_scaled[idx_hier, ]
macro_hier <- spotify$macro_genre[idx_hier]

dist_mat <- dist(X_hier)
hc_fit <- hclust(dist_mat, method = "ward.D2")

hc_cut <- cutree(hc_fit, k = best_k)
ari_hclust <- adjustedRandIndex(hc_cut, as.integer(macro_hier))

tibble(
  Method = c("K-Means", "Hierarchical (Ward)"),
  k = best_k,
  ARI = round(c(ari_kmeans, ari_hclust), 4)
) %>%
  kable(caption = "Adjusted Rand Index: clustering vs. macro-genre labels") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

[PLACEHOLDER: compare k-means vs. hierarchical ARI and interpret. If both ARIs are low (~0–0.15), conclude that genre labels do not map to audio feature clusters — a key finding that motivates the rest of the analysis.]

---

### Angle 2: Within-Genre Variance (Coherence Ranking)

```{r genre-coherence, fig.cap="Top 15 most coherent and bottom 15 least coherent fine-grained genres, ranked by inverse mean within-genre variance of standardized audio features."}
# Compute within-genre variance for all 125 fine-grained genres
coherence_df <- as_tibble(X_scaled) %>%
  mutate(track_genre = spotify$track_genre) %>%
  group_by(track_genre) %>%
  summarise(
    n = n(),
    across(all_of(audio_features), var, .names = "var_{.col}"),
    .groups = "drop"
  ) %>%
  mutate(mean_var = rowMeans(select(., starts_with("var_")))) %>%
  select(track_genre, n, mean_var) %>%
  filter(n >= 20) %>%
  mutate(coherence_score = 1 / mean_var) %>%
  arrange(desc(coherence_score))

top15 <- head(coherence_df, 15)
bottom15 <- tail(coherence_df, 15)

bind_rows(
  top15 %>% mutate(group = "Most Coherent (Top 15)"),
  bottom15 %>% mutate(group = "Least Coherent (Bottom 15)")
) %>%
  ggplot(aes(x = reorder(track_genre, coherence_score),
             y = coherence_score, fill = group)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~group, scales = "free_y", ncol = 2) +
  scale_fill_manual(values = c("Most Coherent (Top 15)" = "#2ecc71",
                               "Least Coherent (Bottom 15)" = "#e74c3c")) +
  labs(title = "Genre Coherence Ranking (Fine-Grained Genres)",
       x = "Genre", y = "Coherence Score (1 / Mean Within-Genre Variance)",
       fill = NULL) +
  theme_minimal() +
  theme(legend.position = "none")

# Macro-genre coherence
macro_coherence <- as_tibble(X_scaled) %>%
  mutate(macro_genre = spotify$macro_genre) %>%
  group_by(macro_genre) %>%
  summarise(
    n = n(),
    across(all_of(audio_features), var, .names = "var_{.col}"),
    .groups = "drop"
  ) %>%
  mutate(mean_var = rowMeans(select(., starts_with("var_")))) %>%
  select(macro_genre, n, mean_var) %>%
  mutate(coherence_score = 1 / mean_var) %>%
  arrange(desc(coherence_score))

macro_coherence %>%
  select(macro_genre, n, mean_var, coherence_score) %>%
  mutate(across(where(is.numeric), ~round(., 4))) %>%
  kable(caption = "Coherence scores at the macro-genre level") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

[PLACEHOLDER: discuss which specific genres are tight (e.g., classical, death-metal) vs. diffuse (e.g., pop, indie). Interpret whether high coherence reflects a strong musical tradition, niche production conventions, or small sample size. Compare macro vs. fine-grained coherence.]

### Does Musical Tightness Pay Off? Coherence vs. Popularity

```{r coherence-vs-popularity, fig.cap="Scatter plot of genre coherence score vs. median track popularity. Point size reflects genre track count. Red line is a linear regression fit with 95% confidence band."}
# Median popularity per fine-grained genre
genre_pop <- spotify %>%
  group_by(track_genre) %>%
  summarise(median_popularity = median(popularity), .groups = "drop")

# Join with coherence scores (already filtered to n >= 20)
coherence_pop <- coherence_df %>%
  left_join(genre_pop, by = "track_genre")

# Label outliers: top/bottom 10% coherence and top 10% popularity
label_threshold_coh_hi  <- quantile(coherence_pop$coherence_score, 0.90)
label_threshold_coh_lo  <- quantile(coherence_pop$coherence_score, 0.10)
label_threshold_pop     <- quantile(coherence_pop$median_popularity, 0.90)

coherence_pop <- coherence_pop %>%
  mutate(label = ifelse(
    coherence_score > label_threshold_coh_hi |
    coherence_score < label_threshold_coh_lo |
    median_popularity > label_threshold_pop,
    track_genre, NA
  ))

ggplot(coherence_pop, aes(x = coherence_score, y = median_popularity)) +
  geom_point(aes(size = n), alpha = 0.55, color = "#3498db") +
  geom_smooth(method = "lm", se = TRUE, color = "#e74c3c", linewidth = 1) +
  geom_text_repel(aes(label = label), size = 3, max.overlaps = 15,
                  segment.color = "grey60") +
  scale_size_continuous(range = c(1.5, 7), name = "Track Count") +
  labs(title = "Genre Coherence vs. Median Popularity",
       subtitle = "Does sounding consistent within a genre correlate with commercial success?",
       x = "Coherence Score (1 / Mean Within-Genre Variance)",
       y = "Median Popularity (0–100)") +
  theme_minimal()

# Regression summary
coh_pop_fit <- lm(median_popularity ~ coherence_score, data = coherence_pop)
coh_summary <- summary(coh_pop_fit)
cat(sprintf("R²: %.4f   |   Slope: %.4f   |   p-value: %.4f\n",
            coh_summary$r.squared,
            coef(coh_pop_fit)["coherence_score"],
            coef(coh_summary)["coherence_score", "Pr(>|t|)"]))
```

The regression yields an R² of 0.0049 and a slope of 2.20 (p = 0.46), indicating no meaningful linear relationship between genre coherence and median popularity. Less than 0.5% of the variance in genre popularity is explained by how internally consistent a genre sounds. The positive slope is not statistically distinguishable from zero at any conventional significance level. This means musical tightness confers no commercial advantage: a genre sounding like itself does not predict how popular its tracks will be on the platform. Popularity appears to be driven by factors orthogonal to acoustic consistency — platform promotion, cultural moment, or artist profile. Notably, `classical` and `iranian` score near zero in popularity despite reasonable coherence, while `pop` and `metal` achieve high median popularity while being acoustically diffuse. This null result reinforces the paper's central argument: genre labels function as market categories rather than acoustic descriptors.

---

### Angle 3: Popularity Regression

#### OLS (Global)

```{r regression-global}
# Prepare regression data with macro_genre dummies
reg_data <- as_tibble(X_scaled) %>%
  mutate(popularity = spotify$popularity,
         macro_genre = spotify$macro_genre)

ols_fit <- lm(popularity ~ ., data = reg_data)

# VIF check (car::vif returns a matrix for models with factors; take first column)
vif_vals <- tryCatch(car::vif(ols_fit), error = function(e) NULL)
if (!is.null(vif_vals)) {
  vif_vec <- if (is.matrix(vif_vals)) vif_vals[, 1] else vif_vals
  vif_df <- tibble(feature = names(vif_vec), VIF = vif_vec) %>%
    filter(VIF > 5)
  if (nrow(vif_df) > 0) {
    vif_df %>%
      mutate(VIF = round(VIF, 3)) %>%
      kable(caption = "Features with VIF > 5 (multicollinearity concern)") %>%
      kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
  }
}

# Extract key stats
ols_summary <- summary(ols_fit)
cat(sprintf("Adjusted R²: %.4f\n", ols_summary$adj.r.squared))

# Coefficient table (top 20 by |t-value|)
coef_df <- as_tibble(coef(ols_summary), rownames = "term") %>%
  rename(estimate = Estimate, se = `Std. Error`,
         t_val = `t value`, p_val = `Pr(>|t|)`) %>%
  mutate(significant = p_val < 0.05) %>%
  arrange(desc(abs(t_val))) %>%
  head(20)

coef_df %>%
  mutate(across(where(is.numeric), ~round(., 4))) %>%
  kable(caption = "OLS coefficients (top 20 by |t-value|)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

[PLACEHOLDER: interpret adjusted R², note which audio features and macro-genre dummies are most significant, and discuss multicollinearity between energy and loudness.]

#### Lasso Regression

```{r regression-lasso, fig.cap="Lasso coefficient path across log(lambda). Vertical lines indicate lambda.min (left) and lambda.1se (right)."}
set.seed(42)

# Build model matrix with macro_genre dummies
X_lasso <- model.matrix(popularity ~ ., data = reg_data)[, -1]
y_lasso <- reg_data$popularity

cv_lasso <- cv.glmnet(X_lasso, y_lasso, alpha = 1, nfolds = 5)

plot(cv_lasso$glmnet.fit, xvar = "lambda", label = FALSE)
abline(v = log(cv_lasso$lambda.min), col = "red", lty = 2)
abline(v = log(cv_lasso$lambda.1se), col = "blue", lty = 2)
legend("topright", legend = c("lambda.min", "lambda.1se"),
       col = c("red", "blue"), lty = 2, bty = "n")
title("Lasso Coefficient Path")

# Surviving features at lambda.1se
lasso_coefs <- coef(cv_lasso, s = "lambda.1se")
lasso_coef_df <- as_tibble(as.matrix(lasso_coefs), rownames = "feature") %>%
  rename(coefficient = s1) %>%
  filter(feature != "(Intercept)", coefficient != 0) %>%
  arrange(desc(abs(coefficient)))

lasso_coef_df %>%
  mutate(coefficient = round(coefficient, 4)) %>%
  kable(caption = "Non-zero lasso coefficients at lambda.1se") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

[PLACEHOLDER: report how many features survive at lambda.1se vs. lambda.min; note whether macro_genre dummies are retained; interpret the direction and magnitude of key surviving coefficients.]

#### Stratified OLS by Macro-Genre

```{r regression-stratified, fig.cap="Comparison of OLS coefficients for key audio features across the 5 largest macro-genres, with 95% confidence intervals."}
# Fit separate OLS models for top 5 macro-genres by track count
top5_genres <- names(sort(table(spotify$macro_genre), decreasing = TRUE))[1:5]

stratified_coefs <- map_dfr(top5_genres, function(g) {
  df_g <- reg_data %>% filter(macro_genre == g) %>% select(-macro_genre)
  fit_g <- lm(popularity ~ ., data = df_g)
  s <- summary(fit_g)$coefficients
  as_tibble(s, rownames = "term") %>%
    rename(estimate = Estimate, se = `Std. Error`) %>%
    mutate(macro_genre = g,
           lower = estimate - 1.96 * se,
           upper = estimate + 1.96 * se)
})

# Focus on the 10 audio features
stratified_coefs %>%
  filter(term %in% audio_features) %>%
  ggplot(aes(x = estimate, y = reorder(term, estimate),
             color = macro_genre, group = macro_genre)) +
  geom_point(position = position_dodge(0.5), size = 2.5) +
  geom_errorbarh(aes(xmin = lower, xmax = upper),
                 position = position_dodge(0.5), height = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Stratified OLS Coefficients by Macro-Genre",
       x = "Coefficient Estimate (±95% CI)",
       y = "Audio Feature", color = "Macro-Genre") +
  theme_minimal()
```

[PLACEHOLDER: discuss whether the effect of valence or energy on popularity differs across macro-genres; interpret sign reversals as evidence that genre context moderates how audio features drive streaming success.]

---

### Angle 4: Boundary Song Detection

#### Random Forest Classifier

```{r classifier-fit}
set.seed(42)

# Prepare classification data
clf_data <- as_tibble(X_scaled) %>%
  mutate(macro_genre = spotify$macro_genre)

# 80/20 stratified split
train_idx <- createDataPartition(clf_data$macro_genre, p = 0.8, list = FALSE)
train_df <- clf_data[train_idx, ]
test_df  <- clf_data[-train_idx, ]

set.seed(42)
rf_fit <- randomForest(
  macro_genre ~ .,
  data = train_df,
  ntree = 300,
  importance = TRUE
)

# Predictions on test set
rf_pred_class <- predict(rf_fit, test_df)
rf_pred_prob  <- predict(rf_fit, test_df, type = "prob")

# Confusion matrix
cm <- confusionMatrix(rf_pred_class, test_df$macro_genre)
cat(sprintf("Overall Accuracy: %.4f\n", cm$overall["Accuracy"]))

# Macro-averaged F1
class_stats <- cm$byClass
macro_f1 <- mean(class_stats[, "F1"], na.rm = TRUE)
cat(sprintf("Macro-averaged F1: %.4f\n", macro_f1))

# Per-class summary table
as_tibble(class_stats[, c("Precision", "Recall", "F1")], rownames = "Class") %>%
  mutate(across(where(is.numeric), ~round(., 4))) %>%
  kable(caption = "Per-class precision, recall, and F1 on test set") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

[PLACEHOLDER: interpret overall accuracy and macro F1. Note which macro-genres are hardest to classify (low F1) and whether this aligns with the coherence ranking — diffuse genres should be harder to classify correctly.]

#### Boundary Songs

```{r boundary-songs}
# Extract boundary songs: true genre predicted with max prob < 0.4
test_meta <- spotify[-train_idx, ] %>%
  select(track_name, artists, macro_genre)

boundary_df <- as_tibble(rf_pred_prob) %>%
  mutate(
    true_genre    = test_df$macro_genre,
    pred_genre    = rf_pred_class,
    max_prob      = apply(rf_pred_prob, 1, max),
    track_name    = test_meta$track_name,
    artists       = test_meta$artists
  ) %>%
  filter(max_prob < 0.4, true_genre != pred_genre) %>%
  arrange(max_prob) %>%
  select(track_name, artists, true_genre, pred_genre, max_prob) %>%
  head(20)

boundary_df %>%
  mutate(max_prob = round(max_prob, 4)) %>%
  kable(caption = "Top 20 boundary songs: misclassified with low confidence") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

[PLACEHOLDER: surface 3–5 specific well-known songs from this list and offer musical interpretations. Example: a hip-hop track with heavy acoustic guitar might be labeled Hip-Hop/R&B but predicted as Folk/Country/Acoustic. These are the most engaging concrete examples in the paper.]

---

### Angle 5: Feature Importance

```{r feature-importance, fig.cap="Variable importance from the random forest (Mean Decrease in Gini), compared with lasso coefficient magnitudes from Angle 3. Features that matter for genre classification may differ from those that drive popularity."}
# Random forest importance
rf_imp <- importance(rf_fit)
rf_imp_df <- as_tibble(rf_imp, rownames = "feature") %>%
  arrange(desc(MeanDecreaseGini))

p_rf_imp <- ggplot(rf_imp_df,
                   aes(x = reorder(feature, MeanDecreaseGini),
                       y = MeanDecreaseGini)) +
  geom_col(fill = "#9b59b6", alpha = 0.85) +
  coord_flip() +
  labs(title = "RF Feature Importance (Genre Classification)",
       x = "Feature", y = "Mean Decrease in Gini") +
  theme_minimal()

# Lasso importance (absolute coefficients, audio features only)
lasso_imp_df <- lasso_coef_df %>%
  filter(feature %in% audio_features) %>%
  mutate(abs_coef = abs(coefficient))

p_lasso_imp <- ggplot(lasso_imp_df,
                      aes(x = reorder(feature, abs_coef),
                          y = abs_coef)) +
  geom_col(fill = "#e67e22", alpha = 0.85) +
  coord_flip() +
  labs(title = "Lasso Feature Importance (Popularity Regression)",
       x = "Feature", y = "|Lasso Coefficient| at lambda.1se") +
  theme_minimal()

gridExtra::grid.arrange(p_rf_imp, p_lasso_imp, ncol = 2)

# Summary table
rf_imp_df %>%
  select(feature, MeanDecreaseGini) %>%
  left_join(lasso_imp_df %>% select(feature, abs_coef), by = "feature") %>%
  mutate(across(where(is.numeric), ~round(., 4))) %>%
  rename(`RF Gini Importance` = MeanDecreaseGini,
         `|Lasso Coef|` = abs_coef) %>%
  kable(caption = "Feature importance: genre classification vs. popularity regression") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

[PLACEHOLDER: identify whether the top genre-classification features (e.g., instrumentalness, acousticness) are the same as the top popularity-prediction features. Divergence here suggests that the dimensions of sound that make genres distinct are not the same dimensions that make tracks popular.]

---

### Angle 6: Audio Feature Geometry — A Word Embedding Analogy

Angles 1–5 establish that genre labels do not form clean clusters in audio space. This angle asks a complementary question: even if genres don't cluster, does the audio feature space possess meaningful geometric *directionality* — analogous to the celebrated word2vec finding that king − man + woman ≈ queen? We define three interpretable axes from pairs of maximally distinct genre centroids and project all genres onto them.

- **Electronic ↔ Acoustic**: `edm` − `acoustic` (dominant driver: acousticness, energy, tempo)
- **Vocal ↔ Instrumental**: `hip-hop` − `classical` (dominant driver: speechiness, instrumentalness)
- **Heavy/Loud ↔ Soft/Ambient**: `death-metal` − `sleep` (dominant driver: loudness, energy, valence)

```{r genre-compass, fig.cap="Top: 2D genre compass projecting all genres onto the Electronic↔Acoustic (x) and Vocal↔Instrumental (y) axes. Point size reflects track count. Bottom: Genre rankings on the Heavy/Loud ↔ Soft/Ambient axis (top and bottom 10)."}
# Genre centroids in standardized feature space (genres with n >= 20)
centroid_tbl <- as_tibble(X_scaled) %>%
  mutate(track_genre = spotify$track_genre) %>%
  group_by(track_genre) %>%
  summarise(n_tracks = n(), across(all_of(audio_features), mean), .groups = "drop") %>%
  filter(n_tracks >= 20, track_genre %in% names(macro_genre_map))

centroid_mat <- centroid_tbl %>% select(all_of(audio_features)) %>% as.matrix()
rownames(centroid_mat) <- centroid_tbl$track_genre

# Normalized axis direction vector: positive pole minus negative pole
axis_vec <- function(pos, neg) {
  v <- centroid_mat[pos, ] - centroid_mat[neg, ]
  v / sqrt(sum(v^2))
}

ax_electronic <- axis_vec("edm",         "acoustic")    # Electronic ↔ Acoustic
ax_vocal      <- axis_vec("hip-hop",     "classical")   # Vocal ↔ Instrumental
ax_intensity  <- axis_vec("death-metal", "sleep")       # Heavy ↔ Soft

# Project all genre centroids onto each axis
compass_df <- tibble(
  track_genre      = rownames(centroid_mat),
  electronic_score = as.numeric(centroid_mat %*% ax_electronic),
  vocal_score      = as.numeric(centroid_mat %*% ax_vocal),
  intensity_score  = as.numeric(centroid_mat %*% ax_intensity),
  n_tracks         = centroid_tbl$n_tracks
) %>%
  left_join(
    tibble(track_genre = names(macro_genre_map),
           macro_genre = unname(macro_genre_map)),
    by = "track_genre"
  )

# 2D Genre Compass
x_lim <- max(abs(compass_df$electronic_score)) * 0.78
y_lim <- max(abs(compass_df$vocal_score))      * 0.78

ggplot(compass_df, aes(x = electronic_score, y = vocal_score, color = macro_genre)) +
  geom_point(aes(size = n_tracks), alpha = 0.75) +
  geom_text_repel(aes(label = track_genre), size = 2.4, max.overlaps = 25,
                  segment.color = "grey70", segment.size = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.4) +
  annotate("label", x =  x_lim, y =  y_lim, label = "Electronic\n+ Vocal",
           size = 2.8, fill = "white", color = "grey30", label.size = 0.3) +
  annotate("label", x = -x_lim, y =  y_lim, label = "Acoustic\n+ Vocal",
           size = 2.8, fill = "white", color = "grey30", label.size = 0.3) +
  annotate("label", x = -x_lim, y = -y_lim, label = "Acoustic\n+ Instrumental",
           size = 2.8, fill = "white", color = "grey30", label.size = 0.3) +
  annotate("label", x =  x_lim, y = -y_lim, label = "Electronic\n+ Instrumental",
           size = 2.8, fill = "white", color = "grey30", label.size = 0.3) +
  scale_color_brewer(palette = "Set3") +
  scale_size_continuous(range = c(1.5, 6), guide = "none") +
  labs(
    title    = "Genre Compass: Audio Feature Axes",
    subtitle = "X: EDM (electronic) ↔ Acoustic  |  Y: Hip-Hop (vocal) ↔ Classical (instrumental)",
    x        = "← Acoustic   |   Electronic →",
    y        = "← Instrumental   |   Vocal →",
    color    = "Macro-Genre"
  ) +
  theme_minimal()

# Intensity axis bar chart (top 10 heavy, top 10 soft)
bind_rows(
  compass_df %>% slice_max(intensity_score, n = 10) %>% mutate(pole = "Heavy/Loud"),
  compass_df %>% slice_min(intensity_score, n = 10) %>% mutate(pole = "Soft/Ambient")
) %>%
  mutate(track_genre = fct_reorder(track_genre, intensity_score)) %>%
  ggplot(aes(x = track_genre, y = intensity_score, fill = pole)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("Heavy/Loud" = "#e74c3c", "Soft/Ambient" = "#3498db")) +
  labs(title = "Intensity Axis: Death-Metal ↔ Sleep",
       subtitle = "Top and bottom 10 genres projected onto the Heavy/Loud → Soft/Ambient direction",
       x = NULL, y = "Projection Score", fill = NULL) +
  theme_minimal()
```

The genre compass produces strikingly interpretable results. The Acoustic+Instrumental quadrant is anchored exactly as expected by classical, sleep, new-age, piano, and opera — the genres with the most distinctive acoustic and instrumental profiles. The opposite corner (Electronic+Vocal) is occupied by latin, reggae, reggaeton, and funk, all of which are energetic, rhythm-driven, and vocal-forward. Singer-songwriter and acoustic sit in the Acoustic+Vocal quadrant, while minimal-techno and detroit-techno appear in the Electronic+Instrumental corner.

A subtler structural finding emerges from the scatter: genres fall along a diagonal rather than filling all four quadrants evenly. Acoustic genres tend to cluster toward the instrumental end, and electronic genres tend toward the vocal end. This correlation — that acoustic production tends to accompany melodic/instrumental composition, while electronic production tends to accompany dance-oriented vocal arrangements — is a genuine geometric property of the audio feature space, not an artifact of our axis choices.

The intensity axis is similarly well-ordered on the soft end: sleep, classical, new-age, piano, and opera rank as the least intense genres by a wide margin. The heavy end produces one notable surprise: drum-and-bass scores higher than death-metal, the axis anchor itself, indicating that drum-and-bass is even more extreme on the composite loudness-energy-valence direction than the genre used to define "heavy." Reggaeton appearing as the second most intense genre — above metalcore and metal — reflects its production style of maximally compressed, loud rhythmic tracks.

Together, these results support a key nuance: while Spotify's genre labels do not form natural clusters in audio space (as shown by the low clustering ARI and the UMAP blob), the space does possess meaningful geometric directionality. The word2vec analogy holds at the axis level — you can define interpretable audio directions and genres sort sensibly along them — even though it fails at the cluster level. Genre labels are not acoustically coherent regions, but they are acoustically ordered.

---

## Discussion and Conclusion

The central question of this project is whether Spotify's genre labels reflect genuine musical coherence — whether songs that share a genre label actually sound alike in audio feature space. Across six analytical angles, the answer is a consistent and qualified no: genre labels do not correspond to natural acoustic groupings, but the audio feature space is not structureless either. The picture that emerges is more nuanced than simple genre meaninglessness.

**Clustering (Angle 1).** Both k-means (ARI = [XX]) and hierarchical clustering (ARI = [XX]) against macro-genre labels produce adjusted Rand indices close to zero, indicating near-random alignment between unsupervised audio clusters and genre assignments. This finding is reinforced by the UMAP projection, which shows nearly all macro-genres collapsing into a single undifferentiated mass — even under the most powerful non-linear dimensionality reduction available. The single exception is Classical/Jazz, which separates visually in both PCA and UMAP space, reflecting its uniquely distinctive audio fingerprint. The implication is direct: Spotify's genre taxonomy does not arise from natural groupings in the 10-dimensional audio feature space.

**Coherence Ranking (Angle 2).** While genres do not cluster globally, they vary dramatically in internal consistency. [Top coherent genres — e.g., classical, death-metal] rank as the most musically tight, while [bottom coherent genres — e.g., pop, indie] are the most diffuse. This variation reveals that "genre" is an operationally inconsistent construct: some labels are precise musical descriptors with strong production conventions, while others function as broad marketing umbrellas that aggregate acoustically heterogeneous tracks. The coherence-popularity analysis finds no relationship between a genre's musical tightness and its median commercial performance (R² = 0.0049, p = 0.46), confirming that popularity is driven by factors orthogonal to acoustic consistency.

**Popularity Regression (Angle 3).** The global OLS model yields an adjusted R² of [XX], indicating that audio features and genre labels together explain only a modest fraction of variance in popularity. [Note whether macro-genre dummies survive lasso regularization at lambda.1se — if they do, genre carries market value beyond audio features; if not, the audio features absorb genre's effect.] The stratified OLS models reveal that the relationship between individual audio features and popularity differs across macro-genres, suggesting that the same sonic quality (e.g., high valence) can be commercially rewarded in one genre context and penalized in another.

**Boundary Songs (Angle 4).** The random forest classifier achieves an overall accuracy of [XX]% and a macro-averaged F1 of [XX], with [easiest macro-genre] as the most classifiable and [hardest macro-genre] as the least. Tracks with maximum predicted probability below 0.4 for any genre — the boundary songs — are concentrated at genre intersections that reflect real musical hybridization. [Name 2–3 specific songs and their true vs. predicted genre pairs.] These cases demonstrate that musical reality is more continuous than categorical labels suggest: some tracks genuinely inhabit the acoustic space between genres, and the classifier correctly identifies them as such.

**Feature Importance (Angle 5).** The random forest assigns highest importance to [top RF features — e.g., instrumentalness, acousticness, tempo] for genre classification, while the lasso retains [top lasso features] as the strongest predictors of popularity. [Note whether the top genre-classification features overlap with the top popularity-prediction features.] Divergence here would suggest that the audio dimensions that make genres acoustically distinct are not the same dimensions that drive streaming success — a finding consistent with genre labels serving a social and organizational function decoupled from commercial audio optimization.

**Audio Feature Geometry (Angle 6).** Despite the absence of clean cluster separation, the audio feature space possesses meaningful geometric directionality. Projecting genre centroids onto three curated axes — Electronic↔Acoustic, Vocal↔Instrumental, and Heavy↔Soft — produces strikingly intuitive orderings. Classical, piano, and opera anchor the Acoustic+Instrumental corner; latin, reggaeton, and reggae cluster in the Electronic+Vocal corner; sleep and new-age sit at the softest extreme of the intensity axis. Genres sort sensibly along all three dimensions, and the word2vec-style analogy holds at the axis level even when it fails at the cluster level. This is the key nuance of the paper: the audio feature space is geometrically structured and directionally interpretable, but it is continuous rather than categorical. Genre boundaries are not acoustically coherent regions — they are arbitrary thresholds imposed on a smooth underlying space.

**Conclusion.** Spotify's genre labels are not meaningful descriptors of musical identity in the acoustic sense. They do not delineate regions of audio feature space, do not predict which tracks sound similar to each other, and do not confer commercial advantage through acoustic consistency. What they do reflect is a set of culturally and commercially determined categories that approximate rough positions on interpretable audio dimensions — acoustic vs. electronic, vocal vs. instrumental, intense vs. ambient — without cleanly partitioning the space. For streaming platforms, this suggests that audio-feature-based recommendation systems could meaningfully outperform genre-based filtering for most macro-genres, since the genre label adds little information about what a track actually sounds like relative to its audio feature vector. Genre labels appear to serve a social and discovery function — they communicate cultural identity and listener expectation — rather than an acoustic one.

**Limitations.** The dataset captures a single snapshot in time and does not account for how genre conventions evolve. Genre labels are platform-assigned and may reflect Spotify's editorial and licensing decisions rather than community consensus or artist self-identification. The 10-feature audio space is a proprietary abstraction and does not capture melody, harmony, lyrical content, or production texture — dimensions that listeners likely use to perceive genre. Finally, the macro-genre mapping used here is one of many possible aggregations; results may differ under alternative genre hierarchies.

---

## Statement on AI Use

This project made substantial use of Claude Code (Anthropic, claude-sonnet-4-6) as an AI coding and writing assistant throughout. The following aspects were AI-assisted:

- **Code generation.** All R code chunks — data preprocessing, PCA, UMAP, k-means and hierarchical clustering, coherence ranking, lasso and OLS regression, random forest classifier, boundary song extraction, feature importance visualization, and the genre compass — were written with AI assistance. Code was reviewed and executed by the authors, and errors and edge cases (e.g., `sort` argument compatibility, `reorder()` factor handling) were debugged iteratively in the AI-assisted session.

- **Analytical design.** The inclusion of UMAP, the coherence-vs-popularity regression, and the genre compass (Angle 6) were proposed by the AI assistant during the project. The core five-angle framing was human-defined; the AI suggested extensions and the word2vec analogy as a novel framing for Angle 6.

- **Prose and interpretation.** Section narratives, result interpretations, and the Discussion and Conclusion were drafted by the AI assistant based on actual outputs shared during the session. Authors reviewed and edited the prose for accuracy and tone.

- **Human-authored components.** The research question, dataset selection, macro-genre mapping, and overall project framing were defined by the authors. All final decisions about analysis scope, interpretation emphasis, and submission content were made by the authors.

No AI tools were used to generate or fabricate data, results, or numerical outputs. All figures and statistics reported in this document were produced by executing the R code against the actual Spotify dataset.

---

## Appendix

### Appendix A: Full Macro-Genre Mapping

```{r appendix-mapping}
tibble(
  fine_genre = names(macro_genre_map),
  macro_genre = unname(macro_genre_map)
) %>%
  arrange(macro_genre, fine_genre) %>%
  kable(caption = "Full mapping from fine-grained genres to macro-genre families") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  scroll_box(height = "400px")
```

### Appendix B: Session Info

```{r appendix-session}
sessionInfo()
```
