---
title: "Modern Data Mining, HW 2"
author:
- Group Member 1
- Group Member 2
- Group Member 3
date: 'Due: 11:59 PM,  Sunday, 02/22'
output:
  html_document:
    code_folding: show
    highlight: haddock
    number_sections: yes
    theme: lumen
    toc: yes
    toc_depth: 4
    toc_float: yes
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: '4'
  word_document:
    toc: yes
    toc_depth: '4'
urlcolor: blue
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width=8, fig.height=4)
options(scipen = 0, digits = 3)  # controls base R output
# check if you have ISLR package, if not, install it
if(!require('pacman')) {install.packages('pacman')}
pacman::p_load(ISLR, tidyverse, data.table, irlba, gridExtra) # add the packages needed!!!
```


\pagebreak

# Overview {-}

Principle Component Analysis is widely used in data exploration, dimension reduction, data visualization. The aim is to transform original data into uncorrelated linear combinations of the original data while keeping the information contained in the data. High dimensional data tends to show clusters in lower dimensional view. 

Clustering Analysis is another form of EDA. Here we are hoping to group data points which are close to each other within the groups and far away between different groups. Clustering using PC's can be effective. Clustering analysis can be very subjective in the way we need to summarize the properties within each group. 

Both PCA and Clustering Analysis are so called unsupervised learning. There is no response variables involved in the process. 

For supervised learning, we try to find out how does a set of predictors relate to some response variable of the interest. Multiple regression is still by far, one of the most popular methods. We use a linear model as a working model for its simplicity and interpretability. It is important that we use domain knowledge as much as we can to determine the form of the response as well as the function format of the factors on the other hand. 



## Objectives

- PCA
- SVD
- Clustering Analysis
- Linear Regression

## Review materials

- Study Module 2: PCA
- Study Module 3: Clustering Analysis
- Study Module 4: Multiple regression (Including Simple regression as well)

## Data needed

- `NLSY79.csv`
- `brca_subtype.csv`
- `brca_x_patient.csv`


# Case study 1: Self-esteem 

Self-esteem generally describes a person's overall sense of self-worthiness and personal value. It can play significant role in one's motivation and success throughout the life. Factors that influence self-esteem can be inner thinking, health condition, age, life experiences etc. We will try to identify possible factors in our data that are related to the level of self-esteem. 

In the well-cited National Longitudinal Study of Youth (NLSY79), it follows about 13,000 individuals and numerous individual-year information has been gathered through surveys. The survey data is open to public [here](https://www.nlsinfo.org/investigator/). Among many variables we assembled a subset of variables including personal demographic variables in different years, household environment in 79, ASVAB test Scores in 81 and Self-Esteem scores in 81 and 87 respectively. 

The data is store in `NLSY79.csv`.



Here are the description of variables:

**Personal Demographic Variables**

* Gender: a factor with levels "female" and "male"
* Education05: years of education completed by 2005
* HeightFeet05, HeightInch05: height measurement. For example, a person of 5'10 will be recorded as HeightFeet05=5, HeightInch05=10.
* Weight05: weight in lbs.
* Income87, Income05: total annual income from wages and salary in 2005. 
* Job87 (missing), Job05: job type in 1987 and 2005, including Protective Service Occupations, Food Preparation and Serving Related Occupations, Cleaning and Building Service Occupations, Entertainment Attendants and Related Workers, Funeral Related Occupations, Personal Care and Service Workers, Sales and Related Workers, Office and Administrative Support Workers, Farming, Fishing and Forestry Occupations, Construction Trade and Extraction Workers, Installation, Maintenance and Repairs Workers, Production and Operating Workers, Food Preparation Occupations, Setters, Operators and Tenders,  Transportation and Material Moving Workers
 
 
**Household Environment**
 
* Imagazine: a variable taking on the value 1 if anyone in the respondent’s household regularly read magazines in 1979, otherwise 0
* Inewspaper: a variable taking on the value 1 if anyone in the respondent’s household regularly read newspapers in 1979, otherwise 0
* Ilibrary: a variable taking on the value 1 if anyone in the respondent’s household had a library card in 1979, otherwise 0
* MotherEd: mother’s years of education
* FatherEd: father’s years of education
* FamilyIncome78

**Variables Related to ASVAB test Scores in 1981**

Test | Description
--------- | ------------------------------------------------------
AFQT | percentile score on the AFQT intelligence test in 1981 
Coding | score on the Coding Speed test in 1981
Auto | score on the Automotive and Shop test in 1981
Mechanic | score on the Mechanic test in 1981
Elec | score on the Electronics Information test in 1981
Science | score on the General Science test in 1981
Math | score on the Math test in 1981
Arith | score on the Arithmetic Reasoning test in 1981
Word | score on the Word Knowledge Test in 1981
Parag | score on the Paragraph Comprehension test in 1981
Numer | score on the Numerical Operations test in 1981

**Self-Esteem test 81 and 87**

We have two sets of self-esteem test, one in 1981 and the other in 1987. Each set has same 10 questions. 
They are labeled as `Esteem81` and `Esteem87` respectively followed by the question number.
For example, `Esteem81_1` is Esteem question 1 in 81.

The following 10 questions are answered as 1: strongly agree, 2: agree, 3: disagree, 4: strongly disagree

* Esteem 1: “I am a person of worth”
* Esteem 2: “I have a number of good qualities”
* Esteem 3: “I am inclined to feel like a failure”
* Esteem 4: “I do things as well as others”
* Esteem 5: “I do not have much to be proud of”
* Esteem 6: “I take a positive attitude towards myself and others”
* Esteem 7: “I am satisfied with myself”
* Esteem 8: “I wish I could have more respect for myself”
* Esteem 9: “I feel useless at times”
* Esteem 10: “I think I am no good at all”

## Data preparation

Load the data. Do a quick EDA to get familiar with the data set. Pay attention to the unit of each variable. Are there any missing values? 

```{r quick skim of the data}
# temp <- read.csv('data/NLSY79.csv', header = T, stringsAsFactors = F)
# # missing values? real variables vs. factors? are variable values reasonable?
# str(temp)
# summary(temp)
# levels(as.factor(temp$Job05))
# table(as.factor(temp$Job05))
```


## Self esteem evaluation

**Let concentrate on Esteem scores evaluated in 87. **

0. First do a quick summary over all the `Esteem` variables. Pay attention to missing values, any peculiar numbers etc. How do you fix problems discovered if there is any? Briefly describe what you have done for the data preparation. 

1. Please note that higher scores on Esteem questions 1, 2, 4, 6, and 7 indicate higher self-esteem, whereas higher scores on the remaining questions suggest lower self-esteem. To maintain consistency, consider reversing the scores of certain Esteem questions. For example, if the esteem data is stored in `data.esteem`, you can use the code `data.esteem[, c(1, 2, 4, 6, 7)] <- 5 - data.esteem[, c(1, 2, 4, 6, 7)]` to invert the scores.

2. Write a brief summary with necessary plots about the 10 esteem measurements.

3. Do esteem scores all positively correlated? Report the pairwise correlation table and write a brief summary.

4. PCA on 10 esteem measurements. (centered but no scaling)

    a) Report the PC1 and PC2 loadings. Are they unit vectors? Are they orthogonal? 
  
    b) Are there good interpretations for PC1 and PC2? (If loadings are all negative, take the positive loadings for the ease of interpretation)
  
    c) How is the PC1 score obtained for each subject? Write down the formula.
    
    d) Are PC1 scores and PC2 scores in the data uncorrelated? 
    
    e) Plot PVE (Proportion of Variance Explained) and summarize the plot. 
  
    f) Also plot CPVE (Cumulative Proportion of Variance Explained). What proportion of the variance in the data is explained by the first two principal components?
  
    g) PC’s provide us with a low dimensional view of the self-esteem scores. Use a biplot with the first two PC's to display the data.  Give an interpretation of PC1 and PC2 from the plot. 

5. Apply k-means to cluster subjects on the original esteem scores

    a) Find a reasonable number of clusters using within sum of squared with elbow rules.
    
    b) Can you summarize common features within each cluster?
    
    c) Can you visualize the clusters with somewhat clear boundaries? You may try different pairs of variables and different PC pairs of the esteem scores.


6. We now try to find out what factors are related to self-esteem? PC1 of all the Esteem scores is a good variable to summarize one's esteem scores. We take PC1 as our response variable. 

    a) Prepare possible factors/variables:
    
      - EDA the data set first. 

      - Personal information: gender, education (05), log(income) in 87, job type in 87.  One way to summarize one's weight and height is via Body Mass Index which is defined as the body mass divided by the square of the body height, and is universally expressed in units of kg/m². Note, you need to create BMI first. Then may include it as one possible predictor. 
          
      - Household environment: Imagazine, Inewspaper, Ilibrary, MotherEd, FatherEd, FamilyIncome78. Do set indicators `Imagazine`, `Inewspaper` and `Ilibrary` as factors. 
    
      - You may use PC1 of ASVAB as level of intelligence
        
    b)   Run a few regression models between PC1 of all the esteem scores in 87 and suitable variables listed in a). Find a final best model with your **own clearly defined criterion**. 

      - How did you land this model? Run a model diagnosis to see if the linear model assumptions are reasonably met. 
        
      - Write a summary of your findings. In particular, explain what and how the variables in the model affect one's self-esteem. 
        


# Case study 2: Breast cancer sub-type


[The Cancer Genome Atlas (TCGA)](https://www.cancer.gov/about-nci/organization/ccg/research/structural-genomics/tcga), a landmark cancer genomics program by National Cancer Institute (NCI), molecularly characterized over 20,000 primary cancer and matched normal samples spanning 33 cancer types. The genome data is open to public from the [Genomic Data Commons Data Portal (GDC)](https://portal.gdc.cancer.gov/).
 
In this study, we focus on 4 sub-types of breast cancer (BRCA): basal-like (basal), Luminal A-like (lumA), Luminal B-like (lumB), HER2-enriched. The sub-type is based on PAM50, a clinical-grade luminal-basal classifier. (We had hoped to download the data for control groups for each type of the cancer. But failed to do so. Please let us know if you find the appropriate data.)

* Luminal A cancers are low-grade, tend to grow slowly and have the best prognosis.
* Luminal B cancers generally grow slightly faster than luminal A cancers and their prognosis is slightly worse.
* HER2-enriched cancers tend to grow faster than luminal cancers and can have a worse prognosis, but they are often successfully treated with targeted therapies aimed at the HER2 protein. 
* Basal-like breast cancers or triple negative breast cancers do not have the three receptors that the other sub-types have so have fewer treatment options.

We will try to use mRNA expression data alone without the labels to classify 4 sub-types. Classification without labels or prediction without outcomes is called unsupervised learning. We will use K-means and spectrum clustering to cluster the mRNA data and see whether the sub-type can be separated through mRNA data.

We first read the data using `data.table::fread()` which is a faster way to read in big data than `read.csv()`. 

```{r}
brca <- fread("data/brca_subtype.csv")

# get the sub-type information
brca_subtype <- brca$BRCA_Subtype_PAM50
brca <- brca[,-1]
```

1. Summary and transformation

    a) How many patients are there in each sub-type? 
    
```{r results='hide'}
# Displays subtype counts
table(brca_subtype)
```
    
There are 208 Basal, 91 Her2, 628 LumA, and 233 LumB patients respectively.

    b) Randomly pick 5 genes and plot the histogram by each sub-type. 
    
```{r fig.width=10, fig.height=8}
# Randomly pick 5 genes
set.seed(42)
genes_5 <- sample(colnames(brca), 5)

# Extract columns and join back with subtype
brca_5 <- brca[, ..genes_5]
brca_5$subtype <- brca_subtype

# Pivot to long format for plotting
brca_long <- pivot_longer(brca_5, cols = -subtype, 
                          names_to = "gene", values_to = "expression")

# Plot the histograms by each sub-type
ggplot(brca_long, aes(x = expression, fill = subtype)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30) +
  facet_grid(gene ~ subtype, scales = "free") +
  labs(title = "mRNA Expression by Gene and Subtype",
       x = "Expression", y = "Density") +
  theme_bw() + 
  theme(strip.text.y = element_text(size = 7, angle = 0))
```
    
We use density instead of count on the y-axis so that histograms are comparable across subtypes, since the subtype    groups are imbalanced 
    

    c) Clean and transform the mRNA sequences by first remove gene with zero count and no variability and then apply logarithmic transform. 
    
```{r}
# remove zero count genes
brca_filtered <- brca[, colSums(abs(brca)) != 0, with=FALSE]

# remove zero variability genes
brca_filtered <- brca_filtered[, apply(brca_filtered, 2, var) != 0, with=FALSE]

# log2 transform
brca_log <- log2(as.matrix(brca_filtered + 1e-10))
```
    


2. Apply kmeans on the transformed dataset with 4 centers (4 clusters) and output the discrepancy table between the real sub-type `brca_subtype` and the cluster labels.

```{r cache=TRUE}
# Run kmeans with 4 centers
set.seed(42)
km <- kmeans(brca_log, centers = 4, nstart = 20)

# output discrepancy table and save for later comparison
km_table_original <- table(km$cluster, brca_subtype)
km_table_original
```


3. Spectrum clustering: to scale or not to scale?

    a) Apply PCA on the centered and scaled dataset. How many PCs should we use and why? You are encouraged to use `irlba::irlba()`. **In order to do so please review the section about SVD in PCA module.** 
    
```{r}

# Center and Scale data
brca_scaled <- scale(brca_log, center = TRUE, scale = TRUE)

# Run PCA with and calculate the leading 50
pca_result <- irlba::irlba(brca_scaled, nv = 50)

# Compute variance explained
var_explained <- pca_result$d^2 / sum(pca_result$d^2)
cum_var <- cumsum(var_explained)

# scree plot
plot(cum_var, type = "b", 
     xlab = "Number of PCs", 
     ylab = "Cumulative Variance Explained",
     main = "PCA Scree Plot")
abline(h = 0.9, col = "red", lty = 2)  # 90% threshold line

invisible(which(cum_var >= 0.90)[1]) # Exact number of PCs for 90% threshold
```
Initially, we calculated the leading 50 PCs from PCA using the irlba() function. After plotting a scree plot from the     variance explained by these 50 PCs, we decided to use 32 PCs as this was the threshold for 90% cumulative variance       explained. We chose this threshold as it captures the majority of meaningful variance while avoiding noise from the      remaining components.
    
    
    b) Plot PC1 vs PC2 of the centered and scaled data and PC1 vs PC2 of the centered but unscaled data side by side. Should we scale or not scale for clustering process? Why? 
    
```{r}
# Compute unscaled data
brca_centered <- scale(brca_log, center = TRUE, scale = FALSE)
pca_unscaled <- irlba::irlba(brca_centered, nv = 50)

# Extract PC1 and PC2 for both
pc_scaled <- data.frame(
  PC1 = pca_result$u[,1],
  PC2 = pca_result$u[,2],
  subtype = brca_subtype
)

pc_unscaled <- data.frame(
  PC1 = pca_unscaled$u[,1],
  PC2 = pca_unscaled$u[,2],
  subtype = brca_subtype
)

# Plot side by side
p1 <- ggplot(pc_scaled, aes(x=PC1, y=PC2, color=subtype)) +
  geom_point(alpha=0.5) +
  labs(title="Scaled") +
  theme_bw()

p2 <- ggplot(pc_unscaled, aes(x=PC1, y=PC2, color=subtype)) +
  geom_point(alpha=0.5) +
  labs(title="Unscaled") +
  theme_bw()

gridExtra::grid.arrange(p1, p2, ncol=2)
```
From the plots, the unscaled centered data shows better defined clusters with clearer separation, particularly for the Basal subtype. We therefore use unscaled data for clustering. This is justified for two reasons: First, the log2 transform from earlier already normalizes magnitude differences between genes, largely addressing the scale issue that scaling typically corrects for. Second, variance differences between genes in gene expression data can reflect real biological activity. High variance genes are genuinely more active across subtypes and are therefore more informative for clustering, while low variance genes may simply reflect noise. Scaling would wrongly treat these two types of genes as equally important, destroying useful signal.
    

4. Spectrum clustering: center but do not scale the data

    a) Use the first 4 PCs of the centered and unscaled data and apply kmeans. Find a reasonable number of clusters using within sum of squared with the elbow rule.
    
```{r}
# extract first 4 PCs from unscaled PCA
pc4 <- pca_unscaled$u[, 1:4]

# compute WSS for k = 1 to 10
wss <- sapply(1:10, function(k) {
  set.seed(42)
  kmeans(pc4, centers = k, nstart = 20)$tot.withinss
})

# elbow plot
plot(1:10, wss, type = "b",
     xlab = "Number of Clusters",
     ylab = "Within Sum of Squares",
     main = "Elbow Plot")
```
From the elbow plot, the WSS begins to flatten around k=4, suggesting 4 as a reasonable number of clusters. This also aligns with the known 4 breast cancer subtypes in the dataset.    
    
    b) Choose an optimal cluster number and apply kmeans. Compare the real sub-type and the clustering label as follows: Plot scatter plot of PC1 vs PC2. Use point color to indicate the true cancer type and point shape to indicate the clustering label. Plot the kmeans centroids with black dots. Summarize how good is clustering results compared to the real sub-type.
    
```{r}
# run kmeans with k=4
set.seed(42)
km4 <- kmeans(pc4, centers = 4, nstart = 20)

# get centroids projected onto PC1/PC2
centroids <- as.data.frame(km4$centers[, 1:2])
colnames(centroids) <- c("PC1", "PC2")

# build dataframe for plotting
plot_df <- data.frame(
  PC1 = pca_unscaled$u[, 1],
  PC2 = pca_unscaled$u[, 2],
  subtype = as.factor(brca_subtype),
  cluster = as.factor(km4$cluster)
)

# plot
ggplot(plot_df, aes(x = PC1, y = PC2, color = subtype, shape = cluster)) +
  geom_point(alpha = 0.6) +
  geom_point(data = centroids, aes(x = PC1, y = PC2), 
             color = "black", size = 3, shape = 16, inherit.aes = FALSE) +
  labs(title = "PC1 vs PC2 with KMeans Clusters",
       color = "True Subtype", shape = "Cluster") +
  theme_bw()
```
Cluster 2 correctly captures the majority of Basal subtype patients and is clearly separated from the other clusters, consistent with Basal's distinct biological profile as triple negative breast cancer. Her2 patients are spread across multiple clusters rather than forming their own, however from the PC plot you can see Her2 points occupy a somewhat distinct region between the Basal cluster and the LumA/LumB group, suggesting that with a larger sample size (currently only 91 patients), kmeans may be able to identify a distinct Her2 cluster. Finally, LumA and LumB are heavily mixed across clusters 1, 3 and 4, as these two subtypes have similar gene expression profiles and are not easily separable even in PC space.
    
    c) Compare the clustering result from applying kmeans to the original data and the clustering result from applying kmeans to 4 PCs. Does PCA help in kmeans clustering? What might be the reasons if PCA helps?
```{r}
cat("Original KMeans:\n")
km_table_original

cat("PCA KMeans (4 PCs):\n")
km_table_pca <- table(km4$cluster, brca_subtype)
km_table_pca
```
Comparing the two discrepancy tables, the clustering results are very similar regardless of whether kmeans is applied to the original high dimensional data or the 4 PCs. This suggests that the fundamental limitation is not dimensionality but rather that LumA and LumB have genuinely similar gene expression profiles that cannot be cleanly separated, and Her2 has too small a sample size to form its own cluster.

That said, PCA still provides practical benefits, reducing from 19669 genes to 4 PCs results in significantly faster computation with no noticeable loss in clustering performance, confirming that the 4 PCs capture the most meaningful variance in the data. This demonstrates that PCA is useful for dimensionality reduction even when it does not dramatically improve clustering quality.+

    d) Now we have an x patient with breast cancer but with unknown sub-type. We have this patient's mRNA sequencing data. Project this x patient to the space of PC1 and PC2. (Hint: remember we remove some gene with no counts or no variablity, take log and centered, then find its PC1 to PC4 scores) Plot this patient in the plot in b) with a black dot as well. Calculate the Euclidean distance between this patient and each of the centroid of the cluster. (Don't forget the clusters are obtained by using 4 PC's) Can you tell which sub-type this patient might have? 
    
```{r}
x_patient <- fread("data/brca_x_patient.csv")

# remove same genes as cleaning
x_filtered <- x_patient[, colnames(brca_filtered), with = FALSE]

# log transform
x_log <- log2(as.matrix(x_filtered + 1e-10))

# center using same means
col_means <- attr(brca_centered, "scaled:center")
x_centered <- scale(x_log, center = col_means, scale = FALSE)

# project onto PCs
x_pc <- x_centered %*% pca_unscaled$v[, 1:4] %*% diag(1/pca_unscaled$d[1:4])
x_df <- data.frame(PC1 = x_pc[1,1], PC2 = x_pc[1,2])

# Add patient to exact same plot as b)
ggplot(plot_df, aes(x = PC1, y = PC2, color = subtype, shape = cluster)) +
  geom_point(alpha = 0.6) +
  geom_point(data = centroids, aes(x = PC1, y = PC2), 
             color = "black", size = 3, shape = 16, inherit.aes = FALSE) +
  geom_point(data = x_df, aes(x = PC1, y = PC2),
             color = "black", size = 2, shape = 16, inherit.aes = FALSE) +
  labs(title = "PC1 vs PC2 with X Patient",
       color = "True Subtype", shape = "Cluster") +
  theme_bw()

# calculate euclidean distances between patient and each centroid
x_pc4 <- data.frame(PC1 = x_pc[1,1], PC2 = x_pc[1,2], 
                     PC3 = x_pc[1,3], PC4 = x_pc[1,4])

# euclidean distance to each centroid
distances <- apply(km4$centers, 1, function(centroid) {
  sqrt(sum((x_pc4 - centroid)^2))
})

cat("Euclidean distances to each cluster centroid:\n")
print(distances)
```
From both the plot and the Euclidean distances, the x patient is clearly closest to the Basal cluster centroid, suggesting this patient has the Basal subtype.

# Case Study: Fuel Efficiency in Automobiles

What determines how fuel efficient a car is? Are Japanese cars more fuel efficient?  To answer thes questions we will build various linear models  using 
 the `Auto` dataset from the book `ISLR`. The original dataset contains information for about 400 different cars built in various years.  To get the data, first install the package ISLR which has been done in the first R-chunk.  The `Auto` dataset should be loaded automatically.  Original data source is here: https://archive.ics.uci.edu/ml/datasets/auto+mpg

Get familiar with this dataset first. Tip: you can use the command `?ISLR::Auto` to view a description of the dataset. Our response variable will me `MPG`: miles per gallon.

## EDA


```{r}
# auto <- ISLR::Auto
# ?ISLR::Auto
# summary(auto) 
# hist(auto$acceleration) # to check if there are some outliers
```
a) Explore the data, list the variables with clear definitions. Set each variable with its appropriate class. For example `origin` should be set as a factor. 

b) How many cars are included in this data set? 

c) EDA, focus on pairwise plots and summary statistics. Briefly summarize your findings and any peculiarities in the data.


## What effect does `time` have on `MPG`?

a) Start with a simple regression of `mpg` vs. `year` and report R's `summary` output. Is `year` a significant variable at the .05 level? State what effect `year` has on `mpg`, if any, according to this model. 

b) Add `horsepower` on top of the variable `year` to your linear model. Is `year` still a significant variable at the .05 level? Give a precise interpretation of the `year`'s effect found here. 

c) The two 95% CI's for the coefficient of year differ among (a) and (b). How would you explain the difference to a non-statistician?

d) Create a model with interaction by fitting `lm(mpg ~ year * horsepower)`. Is the interaction effect significant at .05 level? Explain the year effect (if any). 

## Categorical predictors

Remember that the same variable can play different roles! Take a quick look at the variable `cylinders`, and try to use this variable in the following analyses wisely. We all agree that a larger number of cylinders will lower mpg. However, we can interpret `cylinders` as either a continuous (numeric) variable or a categorical variable.

a) Fit a model that treats `cylinders` as a continuous/numeric variable. Is `cylinders` significant at the 0.01 level? What effect does `cylinders` play in this model?

b) Fit a model that treats `cylinders` as a categorical/factor. Is `cylinders` significant at the .01 level? What is the effect of `cylinders` in this model? Describe the `cylinders` effect over `mpg`. 

c) What are the fundamental differences between treating `cylinders` as a continuous and categorical variable in your models? 

d) Can you test the null hypothesis: fit0: `mpg` is linear in `cylinders` vs. fit1: `mpg` relates to `cylinders` as a categorical variable at .01 level?  


## Results

Final modeling question: we want to explore the effects of each feature as best as possible. You may explore interactions, feature transformations, higher order terms, or other strategies within reason. The model(s) should be as parsimonious (simple) as possible unless the gain in accuracy is significant from your point of view.
  
a) Describe the final model. Include diagnostic plots with particular focus on the model residuals and diagnoses.

b) Summarize the effects found.

c) Predict the `mpg` of the following car: A red car built in the US in 1983 that is 180 inches long, has eight cylinders, displaces 350 cu. inches, weighs 4000 pounds, and has a horsepower of 260. Also give a 95% CI for your prediction.


# Simple Regression through simulations 
    
## Linear model through simulations

This exercise is designed to help you understand the linear model using simulations. In this exercise, we will generate $(x_i, y_i)$ pairs so that all linear model assumptions are met.

Presume that $\mathbf{x}$ and $\mathbf{y}$ are linearly related with a normal error $\boldsymbol{\varepsilon}$ , such that $\mathbf{y} = 1 + 1.2\mathbf{x} + \boldsymbol{\varepsilon}$. The standard deviation of the error $\varepsilon_i$ is $\sigma = 2$. 

We can create a sample input vector ($n = 40$) for $\mathbf{x}$ with the following code:

```{r, eval = F, echo = TRUE}
# Generates a vector of size 40 with equally spaced values between 0 and 1, inclusive
x <- seq(0, 1, length = 40)
```


### Generate data

Create a corresponding output vector for $\mathbf{y}$ according to the equation given above. Use `set.seed(1)`. Then, create a scatterplot with $(x_i, y_i)$ pairs. Base R plotting is acceptable, but if you can, please attempt to use `ggplot2` to create the plot. Make sure to have clear labels and sensible titles on your plots.


### Understand the model
i. Find the LS estimates of $\boldsymbol{\beta}_0$ and $\boldsymbol{\beta}_1$, using the `lm()` function. What are the true values of $\boldsymbol{\beta}_0$ and $\boldsymbol{\beta}_1$? Do the estimates look to be good? 

ii. What is your RSE for this linear model fit? Is it close to $\sigma = 2$? 

ii. What is the 95% confidence interval for $\boldsymbol{\beta}_1$? Does this confidence interval capture the true $\boldsymbol{\beta}_1$?

iii. Overlay the LS estimates and the true lines of the mean function onto a copy of the scatterplot you made above.


### diagnoses

i. Provide residual plot where fitted $\mathbf{y}$-values are on the x-axis and residuals are on the y-axis. 

ii. Provide a normal QQ plot of the residuals.

iii. Comment on how well the model assumptions are met for the sample you used. 



## Understand sampling distribution and confidence intervals

This part aims to help you understand the notion of sampling statistics and confidence intervals. Let's concentrate on estimating the slope only.  

Generate 100 samples of size $n = 40$, and estimate the slope coefficient from each sample. Also construct 95\% confidence intervals for the slope. 

i. Summarize the LS estimates of the slope. Does the sampling distribution agree with theory? (First specify the theoretical sampling distribution of the LS estimate.)

ii.  How many of your 95% confidence intervals capture the true $\boldsymbol{\beta}_1$? Display your confidence intervals graphically. 




