---
title: "Modern Data Mining, HW 2"
author:
- Group Member 1
- Group Member 2
- Group Member 3
date: 'Due: 11:59 PM,  Sunday, 02/22'
output:
  html_document:
    code_folding: show
    highlight: haddock
    number_sections: yes
    theme: lumen
    toc: yes
    toc_depth: 4
    toc_float: yes
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: '4'
  word_document:
    toc: yes
    toc_depth: '4'
urlcolor: blue
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width=8, fig.height=4)
options(scipen = 0, digits = 3)  # controls base R output
# check if you have ISLR package, if not, install it
if(!require('pacman')) {install.packages('pacman')}
pacman::p_load(ISLR, tidyverse, data.table) # add the packages needed!!!
```


\pagebreak

# Overview {-}

Principle Component Analysis is widely used in data exploration, dimension reduction, data visualization. The aim is to transform original data into uncorrelated linear combinations of the original data while keeping the information contained in the data. High dimensional data tends to show clusters in lower dimensional view. 

Clustering Analysis is another form of EDA. Here we are hoping to group data points which are close to each other within the groups and far away between different groups. Clustering using PC's can be effective. Clustering analysis can be very subjective in the way we need to summarize the properties within each group. 

Both PCA and Clustering Analysis are so called unsupervised learning. There is no response variables involved in the process. 

For supervised learning, we try to find out how does a set of predictors relate to some response variable of the interest. Multiple regression is still by far, one of the most popular methods. We use a linear model as a working model for its simplicity and interpretability. It is important that we use domain knowledge as much as we can to determine the form of the response as well as the function format of the factors on the other hand. 



## Objectives

- PCA
- SVD
- Clustering Analysis
- Linear Regression

## Review materials

- Study Module 2: PCA
- Study Module 3: Clustering Analysis
- Study Module 4: Multiple regression (Including Simple regression as well)

## Data needed

- `NLSY79.csv`
- `brca_subtype.csv`
- `brca_x_patient.csv`


# Case study 1: Self-esteem 

Self-esteem generally describes a person's overall sense of self-worthiness and personal value. It can play significant role in one's motivation and success throughout the life. Factors that influence self-esteem can be inner thinking, health condition, age, life experiences etc. We will try to identify possible factors in our data that are related to the level of self-esteem. 

In the well-cited National Longitudinal Study of Youth (NLSY79), it follows about 13,000 individuals and numerous individual-year information has been gathered through surveys. The survey data is open to public [here](https://www.nlsinfo.org/investigator/). Among many variables we assembled a subset of variables including personal demographic variables in different years, household environment in 79, ASVAB test Scores in 81 and Self-Esteem scores in 81 and 87 respectively. 

The data is store in `NLSY79.csv`.



Here are the description of variables:

**Personal Demographic Variables**

* Gender: a factor with levels "female" and "male"
* Education05: years of education completed by 2005
* HeightFeet05, HeightInch05: height measurement. For example, a person of 5'10 will be recorded as HeightFeet05=5, HeightInch05=10.
* Weight05: weight in lbs.
* Income87, Income05: total annual income from wages and salary in 2005. 
* Job87 (missing), Job05: job type in 1987 and 2005, including Protective Service Occupations, Food Preparation and Serving Related Occupations, Cleaning and Building Service Occupations, Entertainment Attendants and Related Workers, Funeral Related Occupations, Personal Care and Service Workers, Sales and Related Workers, Office and Administrative Support Workers, Farming, Fishing and Forestry Occupations, Construction Trade and Extraction Workers, Installation, Maintenance and Repairs Workers, Production and Operating Workers, Food Preparation Occupations, Setters, Operators and Tenders,  Transportation and Material Moving Workers
 
 
**Household Environment**
 
* Imagazine: a variable taking on the value 1 if anyone in the respondent’s household regularly read magazines in 1979, otherwise 0
* Inewspaper: a variable taking on the value 1 if anyone in the respondent’s household regularly read newspapers in 1979, otherwise 0
* Ilibrary: a variable taking on the value 1 if anyone in the respondent’s household had a library card in 1979, otherwise 0
* MotherEd: mother’s years of education
* FatherEd: father’s years of education
* FamilyIncome78

**Variables Related to ASVAB test Scores in 1981**

Test | Description
--------- | ------------------------------------------------------
AFQT | percentile score on the AFQT intelligence test in 1981 
Coding | score on the Coding Speed test in 1981
Auto | score on the Automotive and Shop test in 1981
Mechanic | score on the Mechanic test in 1981
Elec | score on the Electronics Information test in 1981
Science | score on the General Science test in 1981
Math | score on the Math test in 1981
Arith | score on the Arithmetic Reasoning test in 1981
Word | score on the Word Knowledge Test in 1981
Parag | score on the Paragraph Comprehension test in 1981
Numer | score on the Numerical Operations test in 1981

**Self-Esteem test 81 and 87**

We have two sets of self-esteem test, one in 1981 and the other in 1987. Each set has same 10 questions. 
They are labeled as `Esteem81` and `Esteem87` respectively followed by the question number.
For example, `Esteem81_1` is Esteem question 1 in 81.

The following 10 questions are answered as 1: strongly agree, 2: agree, 3: disagree, 4: strongly disagree

* Esteem 1: “I am a person of worth”
* Esteem 2: “I have a number of good qualities”
* Esteem 3: “I am inclined to feel like a failure”
* Esteem 4: “I do things as well as others”
* Esteem 5: “I do not have much to be proud of”
* Esteem 6: “I take a positive attitude towards myself and others”
* Esteem 7: “I am satisfied with myself”
* Esteem 8: “I wish I could have more respect for myself”
* Esteem 9: “I feel useless at times”
* Esteem 10: “I think I am no good at all”

## Data preparation

Load the data. Do a quick EDA to get familiar with the data set. Pay attention to the unit of each variable. Are there any missing values? 

```{r quick skim of the data}
# temp <- read.csv('data/NLSY79.csv', header = T, stringsAsFactors = F)
# # missing values? real variables vs. factors? are variable values reasonable?
# str(temp)
# summary(temp)
# levels(as.factor(temp$Job05))
# table(as.factor(temp$Job05))
```


## Self esteem evaluation

**Let concentrate on Esteem scores evaluated in 87. **

0. First do a quick summary over all the `Esteem` variables. Pay attention to missing values, any peculiar numbers etc. How do you fix problems discovered if there is any? Briefly describe what you have done for the data preparation. 

1. Please note that higher scores on Esteem questions 1, 2, 4, 6, and 7 indicate higher self-esteem, whereas higher scores on the remaining questions suggest lower self-esteem. To maintain consistency, consider reversing the scores of certain Esteem questions. For example, if the esteem data is stored in `data.esteem`, you can use the code `data.esteem[, c(1, 2, 4, 6, 7)] <- 5 - data.esteem[, c(1, 2, 4, 6, 7)]` to invert the scores.

2. Write a brief summary with necessary plots about the 10 esteem measurements.

3. Do esteem scores all positively correlated? Report the pairwise correlation table and write a brief summary.

4. PCA on 10 esteem measurements. (centered but no scaling)

    a) Report the PC1 and PC2 loadings. Are they unit vectors? Are they orthogonal? 
  
    b) Are there good interpretations for PC1 and PC2? (If loadings are all negative, take the positive loadings for the ease of interpretation)
  
    c) How is the PC1 score obtained for each subject? Write down the formula.
    
    d) Are PC1 scores and PC2 scores in the data uncorrelated? 
    
    e) Plot PVE (Proportion of Variance Explained) and summarize the plot. 
  
    f) Also plot CPVE (Cumulative Proportion of Variance Explained). What proportion of the variance in the data is explained by the first two principal components?
  
    g) PC’s provide us with a low dimensional view of the self-esteem scores. Use a biplot with the first two PC's to display the data.  Give an interpretation of PC1 and PC2 from the plot. 

5. Apply k-means to cluster subjects on the original esteem scores

    a) Find a reasonable number of clusters using within sum of squared with elbow rules.
    
    b) Can you summarize common features within each cluster?
    
    c) Can you visualize the clusters with somewhat clear boundaries? You may try different pairs of variables and different PC pairs of the esteem scores.


6. We now try to find out what factors are related to self-esteem? PC1 of all the Esteem scores is a good variable to summarize one's esteem scores. We take PC1 as our response variable. 

    a) Prepare possible factors/variables:
    
      - EDA the data set first. 

      - Personal information: gender, education (05), log(income) in 87, job type in 87.  One way to summarize one's weight and height is via Body Mass Index which is defined as the body mass divided by the square of the body height, and is universally expressed in units of kg/m². Note, you need to create BMI first. Then may include it as one possible predictor. 
          
      - Household environment: Imagazine, Inewspaper, Ilibrary, MotherEd, FatherEd, FamilyIncome78. Do set indicators `Imagazine`, `Inewspaper` and `Ilibrary` as factors. 
    
      - You may use PC1 of ASVAB as level of intelligence
        
    b)   Run a few regression models between PC1 of all the esteem scores in 87 and suitable variables listed in a). Find a final best model with your **own clearly defined criterion**. 

      - How did you land this model? Run a model diagnosis to see if the linear model assumptions are reasonably met. 
        
      - Write a summary of your findings. In particular, explain what and how the variables in the model affect one's self-esteem. 
        


# Case study 2: Breast cancer sub-type


[The Cancer Genome Atlas (TCGA)](https://www.cancer.gov/about-nci/organization/ccg/research/structural-genomics/tcga), a landmark cancer genomics program by National Cancer Institute (NCI), molecularly characterized over 20,000 primary cancer and matched normal samples spanning 33 cancer types. The genome data is open to public from the [Genomic Data Commons Data Portal (GDC)](https://portal.gdc.cancer.gov/).
 
In this study, we focus on 4 sub-types of breast cancer (BRCA): basal-like (basal), Luminal A-like (lumA), Luminal B-like (lumB), HER2-enriched. The sub-type is based on PAM50, a clinical-grade luminal-basal classifier. (We had hoped to download the data for control groups for each type of the cancer. But failed to do so. Please let us know if you find the appropriate data.)

* Luminal A cancers are low-grade, tend to grow slowly and have the best prognosis.
* Luminal B cancers generally grow slightly faster than luminal A cancers and their prognosis is slightly worse.
* HER2-enriched cancers tend to grow faster than luminal cancers and can have a worse prognosis, but they are often successfully treated with targeted therapies aimed at the HER2 protein. 
* Basal-like breast cancers or triple negative breast cancers do not have the three receptors that the other sub-types have so have fewer treatment options.

We will try to use mRNA expression data alone without the labels to classify 4 sub-types. Classification without labels or prediction without outcomes is called unsupervised learning. We will use K-means and spectrum clustering to cluster the mRNA data and see whether the sub-type can be separated through mRNA data.

We first read the data using `data.table::fread()` which is a faster way to read in big data than `read.csv()`. 

```{r}
brca <- fread("data/brca_subtype.csv")

# get the sub-type information
brca_subtype <- brca$BRCA_Subtype_PAM50
brca <- brca[,-1]
```

1. Summary and transformation

    a) How many patients are there in each sub-type? 

    b) Randomly pick 5 genes and plot the histogram by each sub-type. 

    c) Clean and transform the mRNA sequences by first remove gene with zero count and no variability and then apply logarithmic transform. 


2. Apply kmeans on the transformed dataset with 4 centers (4 clusters) and output the discrepancy table between the real sub-type `brca_subtype` and the cluster labels.

3. Spectrum clustering: to scale or not to scale?

    a) Apply PCA on the centered and scaled dataset. How many PCs should we use and why? You are encouraged to use `irlba::irlba()`. **In order to do so please review the section about SVD in PCA module.** 
    
    b) Plot PC1 vs PC2 of the centered and scaled data and PC1 vs PC2 of the centered but unscaled data side by side. Should we scale or not scale for clustering process? Why? 

4. Spectrum clustering: center but do not scale the data

    a) Use the first 4 PCs of the centered and unscaled data and apply kmeans. Find a reasonable number of clusters using within sum of squared with the elbow rule.
    
    b) Choose an optimal cluster number and apply kmeans. Compare the real sub-type and the clustering label as follows: Plot scatter plot of PC1 vs PC2. Use point color to indicate the true cancer type and point shape to indicate the clustering label. Plot the kmeans centroids with black dots. Summarize how good is clustering results compared to the real sub-type.
    
    c) Compare the clustering result from applying kmeans to the original data and the clustering result from applying kmeans to 4 PCs. Does PCA help in kmeans clustering? What might be the reasons if PCA helps?
    
    d) Now we have an x patient with breast cancer but with unknown sub-type. We have this patient's mRNA sequencing data. Project this x patient to the space of PC1 and PC2. (Hint: remember we remove some gene with no counts or no variablity, take log and centered, then find its PC1 to PC4 scores) Plot this patient in the plot in b) with a black dot as well. Calculate the Euclidean distance between this patient and each of the centroid of the cluster. (Don't forget the clusters are obtained by using 4 PC's) Can you tell which sub-type this patient might have? 
    
```{r}
x_patient <- fread("data/brca_x_patient.csv")
```


# Case Study: Fuel Efficiency in Automobiles

What determines how fuel efficient a car is? Are Japanese cars more fuel efficient?  To answer thes questions we will build various linear models  using 
 the `Auto` dataset from the book `ISLR`. The original dataset contains information for about 400 different cars built in various years.  To get the data, first install the package ISLR which has been done in the first R-chunk.  The `Auto` dataset should be loaded automatically.  Original data source is here: https://archive.ics.uci.edu/ml/datasets/auto+mpg

Get familiar with this dataset first. Tip: you can use the command `?ISLR::Auto` to view a description of the dataset. Our response variable will me `MPG`: miles per gallon.

## EDA


```{r}
auto <- ISLR::Auto
?ISLR::Auto
summary(auto) 
hist(auto$acceleration) # to check if there are some outliers
```
a) Explore the data, list the variables with clear definitions. Set each variable with its appropriate class. For example `origin` should be set as a factor. 

```{r}
library(ISLR)
library(tidyverse)

# Show Variable with definition
auto <- ISLR::Auto
sapply(auto, class)
```

```{r}

# Convert variables to appropriate types
auto <- auto %>%
  mutate(
    origin = factor(origin,
                    levels = c(1,2,3),
                    labels = c("USA","Europe","Japan")),
    cylinders = factor(cylinders),   # engine category, not numeric
    name = as.character(name),       # identifier, not a feature
    year = as.integer(year)          # numeric time index
  )
sapply(auto, class)
str(auto)
```


b) How many cars are included in this data set? 

```{r}
nrow(auto)
```


c) EDA, focus on pairwise plots and summary statistics. Briefly summarize your findings and any peculiarities in the data.

```{r}
auto_num <- auto[, c("mpg","displacement","horsepower",
                     "weight","acceleration","year")]

pairs(auto_num)

data.frame(
  Mean = sapply(auto_num, mean),
  SD   = sapply(auto_num, sd),
  Min  = sapply(auto_num, min),
  Max  = sapply(auto_num, max)
)
```
The scatterplot matrix reveals several strong relationships among the variables. Fuel efficiency (mpg) is strongly negatively associated with weight, displacement, and horsepower, indicating that heavier vehicles with larger and more powerful engines tend to have lower fuel efficiency. In contrast, mpg shows a clear positive relationship with model year, suggesting that automobiles became more fuel efficient over time due to technological improvements. Displacement, horsepower, and weight are highly correlated with each other, indicating potential multicollinearity among engine size variables. The summary statistics show a wide range of mpg values but no extreme outliers. The acceleration variable appears approximately symmetric, suggesting no major violations of modeling assumptions.

## What effect does `time` have on `MPG`?

a) Start with a simple regression of `mpg` vs. `year` and report R's `summary` output. Is `year` a significant variable at the .05 level? State what effect `year` has on `mpg`, if any, according to this model. 

```{r}
fit_a <- lm(mpg ~ year, data = auto)
summary(fit_a)
```
Response: Yes, year is statistically significant at the 0.05 level since p << 0.05. For each additional model year (a car that is one year newer), the expected fuel efficiency increases by about 1.23 miles per gallon on average, according to this model.


b) Add `horsepower` on top of the variable `year` to your linear model. Is `year` still a significant variable at the .05 level? Give a precise interpretation of the `year`'s effect found here. 

```{r}
fit_b <- lm(mpg ~ year + horsepower, data = auto)
summary(fit_b)
confint(fit_b, "year", level = 0.95)
```
Response: Yes, year is still statistically significant at the 0.05 level after adding horsepower. For two cars with the same horsepower, a car that is one model year newer is expected to have about 0.66 mpg higher fuel efficiency on average. We are 95% confident that, after controlling for horsepower, each additional model year increases fuel efficiency by between about 0.53 and 0.79 mpg on average. Because the interval does not include 0, the effect is significant. Compare to 3.2.a, the effect of mpg got smaller because in the simple model, year was partly capturing differences in engine power. Once horsepower is controlled for, year represents improvements beyond engine size (i.e. technology, aerodynamics, efficiency).

c) The two 95% CI's for the coefficient of year differ among (a) and (b). How would you explain the difference to a non-statistician?
Response: In model (a), we only used the car’s model year to predict mpg, so the “year effect” is doing double duty where it captures not only improvements in technology over time but also the fact that newer cars tend to have different engine characteristics (including horsepower). In model (b), we separate those effects by comparing cars from different years while holding horsepower the same. Once horsepower is accounted for, the year coefficient measures a more “pure” technology/regulation/time effect, so the estimated year effect becomes smaller and the confidence interval changes. In short: (a) compares “newer vs older cars in general,” while (b) compares “newer vs older cars that have the same horsepower,” so they answer different questions and therefore give different 95% confidence intervals.

d) Create a model with interaction by fitting `lm(mpg ~ year * horsepower)`. Is the interaction effect significant at .05 level? Explain the year effect (if any). 
```{r}
fit_int <- lm(mpg ~ year * horsepower, data = auto)
summary(fit_int)
confint(fit_int, "year:horsepower", level = 0.95)
```
Response: Fitting the interaction model mpg∼year∗horsepower, the interaction term year:horsepower is highly significant at the 0.05 level (p-value < 0.05; 95% CI [−0.019,−0.0129]). This indicates that the effect of model year on mpg depends on horsepower. Specifically, the marginal effect of year is 2.19 − 0.016 * (horsepower), so the positive effect of newer model years on fuel efficiency becomes smaller as horsepower increases. In other words, mpg tends to improve more over time for lower-horsepower cars than for high-horsepower cars.

## Categorical predictors

Remember that the same variable can play different roles! Take a quick look at the variable `cylinders`, and try to use this variable in the following analyses wisely. We all agree that a larger number of cylinders will lower mpg. However, we can interpret `cylinders` as either a continuous (numeric) variable or a categorical variable.

a) Fit a model that treats `cylinders` as a continuous/numeric variable. Is `cylinders` significant at the 0.01 level? What effect does `cylinders` play in this model?

```{r}
# Treat cylinders as numeric just for this model
auto$cyl_num <- as.numeric(as.character(auto$cylinders))

fit_cyl_num <- lm(mpg ~ cyl_num, data = auto)
summary(fit_cyl_num)
confint(fit_cyl_num, "cyl_num", level = 0.99)  # alpha = 0.01
```
Response: Treating cylinders as a numeric variable, the coefficient for cylinders is highly statistically significant at the 0.01 level (p-value < 0.01). The fitted model is mpg = 42.916 - 3.558 * (cylinders). This indicates that, according to the model, each additional cylinder is associated with a decrease of approximately 3.56 miles per gallon in expected fuel efficiency on average. The 99% confidence interval for the coefficient [−3.94,−3.18] does not include 0, confirming statistical significance.

b) Fit a model that treats `cylinders` as a categorical/factor. Is `cylinders` significant at the .01 level? What is the effect of `cylinders` in this model? Describe the `cylinders` effect over `mpg`. 

```{r}
fit_cyl_fac <- lm(mpg ~ cylinders, data = auto)
summary(fit_cyl_fac)
confint(fit_cyl_fac, level = 0.99)
```
Response: When cylinders is treated as a categorical variable, the overall variable is highly significant at the 0.01 level (F-statistic = 173, p-value < 0.01). In this model, cylinders does not impose a linear “per-cylinder” effect; instead, it estimates separate average mpg values for each cylinder group relative to the baseline (3 cylinders). Overall, higher-cylinder cars tend to have lower mpg, with 8-cylinder cars having the lowest predicted fuel efficiency, while 4- and 5-cylinder cars have the highest predicted mpg.

c) What are the fundamental differences between treating `cylinders` as a continuous and categorical variable in your models? 
Response: Treating cylinders as a continuous variable forces the model to assume a linear relationship between cylinders and mpg: every one-cylinder increase is assumed to change mpg by the same fixed amount, so the effect of going from 3→4 cylinders is constrained to be the same as going from 6→7 (even though some counts may not appear). In contrast, treating cylinders as a categorical (factor) variable does not impose linearity but instead, the model estimates a separate average mpg for each cylinder group, so mpg can differ in a step-like, non-uniform way across 3-, 4-, 5-, 6-, and 8-cylinder cars.

d) Can you test the null hypothesis: fit0: `mpg` is linear in `cylinders` vs. fit1: `mpg` relates to `cylinders` as a categorical variable at .01 level?  
```{r}
# numeric version of cylinders
auto$cyl_num <- as.numeric(as.character(auto$cylinders))

# fit0: linear in cylinders
fit0 <- lm(mpg ~ cyl_num, data = auto)

# fit1: categorical cylinders
fit1 <- lm(mpg ~ cylinders, data = auto)

anova(fit0, fit1)
```
Response: We test whether mpg is linear in cylinders by comparing a model that treats cylinders as a numeric variable with a model that treats cylinders as a categorical variable using an ANOVA F-test. The test gives an F value of 13.2 and a p-value of 3.4×10⁻⁸, which is less than 0.01. Therefore, we reject the null hypothesis at the 1% significance level. This means a single straight-line relationship between mpg and cylinders is not adequate, and modeling cylinders as separate categories provides a significantly better fit to the data.


## Results

Final modeling question: we want to explore the effects of each feature as best as possible. You may explore interactions, feature transformations, higher order terms, or other strategies within reason. The model(s) should be as parsimonious (simple) as possible unless the gain in accuracy is significant from your point of view.
  
a) Describe the final model. Include diagnostic plots with particular focus on the model residuals and diagnoses.
```{r}
auto <- ISLR::Auto

auto <- auto %>%
  mutate(
    origin = factor(origin,
                    levels = c(1,2,3),
                    labels = c("USA","Europe","Japan")),
    cylinders = factor(cylinders),
    name = as.character(name)
  )
levels(auto$origin)
levels(auto$cylinders)
table(auto$origin)
table(auto$cylinders)

m0 <- lm(mpg ~ year + weight + horsepower + origin + cylinders, data = auto)
summary(m0)

```

```{r}
m1 <- lm(mpg ~ year * horsepower + weight + origin + cylinders, data = auto)

# Compare nested models
anova(m0, m1)

# Compare fit metrics
AIC(m0, m1)
BIC(m0, m1)
summary(m0)$adj.r.squared
summary(m1)$adj.r.squared
```
So m1 is indeed better than m0. Now we run some diagnostic plots for m1.
```{r}
par(mfrow=c(2,2))
plot(m1)
par(mfrow=c(1,1))

hist(resid(m1), main="Histogram of residuals", xlab="Residual")

cook <- cooks.distance(m1)
which(cook > 4/length(cook))
```

b) Summarize the effects found.
Response: The final model used is a multiple linear regression with mpg predicted by year, horsepower (including their interaction), weight, origin, and cylinders. The interaction between year and horsepower is kept because it significantly improves the model fit (F-test p-value < 0.001, adjusted R² increases, and AIC/BIC decrease), indicating that improvements in fuel efficiency over time depend on engine power. Diagnostic plots were examined to assess the model assumptions. The residuals versus fitted plot shows residuals randomly scattered around zero with no strong curvature, suggesting the linear form is appropriate. The Q–Q plot and histogram show the residuals are approximately normally distributed with only mild deviations in the tails. The scale-location plot indicates roughly constant variance with only slight spread at higher fitted values, suggesting no serious heteroskedasticity. The residuals versus leverage plot shows a few moderately influential observations, but none exceed Cook’s distance thresholds, so no single observation unduly affects the model. Overall, the regression assumptions appear reasonably satisfied and the model provides a good fit to the data.


c) Predict the `mpg` of the following car: A red car built in the US in 1983 that is 180 inches long, has eight cylinders, displaces 350 cu. inches, weighs 4000 pounds, and has a horsepower of 260. Also give a 95% CI for your prediction.
```{r}
# Make sure origin and cylinders levels match training data
new_car <- data.frame(
  year = 83,
  horsepower = 260,
  weight = 4000,
  origin = factor("USA", levels = levels(auto$origin)),
  cylinders = factor("8", levels = levels(auto$cylinders))
)

# Point prediction + 95% prediction interval for a single car
predict(m1, newdata = new_car, interval = "prediction", level = 0.95)
```


# Simple Regression through simulations 
    
## Linear model through simulations

This exercise is designed to help you understand the linear model using simulations. In this exercise, we will generate $(x_i, y_i)$ pairs so that all linear model assumptions are met.

Presume that $\mathbf{x}$ and $\mathbf{y}$ are linearly related with a normal error $\boldsymbol{\varepsilon}$ , such that $\mathbf{y} = 1 + 1.2\mathbf{x} + \boldsymbol{\varepsilon}$. The standard deviation of the error $\varepsilon_i$ is $\sigma = 2$. 

We can create a sample input vector ($n = 40$) for $\mathbf{x}$ with the following code:

```{r, eval = F, echo = TRUE}
# Generates a vector of size 40 with equally spaced values between 0 and 1, inclusive
x <- seq(0, 1, length = 40)
```


### Generate data

Create a corresponding output vector for $\mathbf{y}$ according to the equation given above. Use `set.seed(1)`. Then, create a scatterplot with $(x_i, y_i)$ pairs. Base R plotting is acceptable, but if you can, please attempt to use `ggplot2` to create the plot. Make sure to have clear labels and sensible titles on your plots.

```{r}
set.seed(1)

n <- 40

# generate normal errors with sd = 2
eps <- rnorm(n, mean = 0, sd = 2)

# generate y from the true model
y <- 1 + 1.2*x + eps

sim_data <- data.frame(x = x, y = y)

library(ggplot2)

ggplot(sim_data, aes(x = x, y = y)) +
  geom_point(color = "steelblue", size = 2) +
  labs(
    title = "Simulated Data from Linear Model",
    subtitle = "y = 1 + 1.2x + ε,  ε ~ N(0, 2²)",
    x = "x",
    y = "y"
  ) +
  theme_minimal()
```

### Understand the model
i. Find the LS estimates of $\boldsymbol{\beta}_0$ and $\boldsymbol{\beta}_1$, using the `lm()` function. What are the true values of $\boldsymbol{\beta}_0$ and $\boldsymbol{\beta}_1$? Do the estimates look to be good? 
```{r}
fit <- lm(y ~ x, data = sim_data)
summary(fit)
coef(fit)
```
Response: The true values used to generate the data were β0 = 1 and β1 = 1.2. Estimated values should be close to these but not exactly equal because the data include random noise with standard deviation 2 and the sample size is only 40. The estimates look good if the intercept is around 1 and the slope is around 1.2 and the fitted line matches the trend in the scatterplot.

ii. What is your RSE for this linear model fit? Is it close to $\sigma = 2$? 
Response: RSE = 1.79. The true error standard deviation used to generate the data was σ = 2. The RSE is an estimate of this quantity, so it should be close but not identical because it is computed from a random sample of size 40. Since 1.79 is reasonably close to 2, the fitted model is doing a good job estimating the true noise level in the data.

ii. What is the 95% confidence interval for $\boldsymbol{\beta}_1$? Does this confidence interval capture the true $\boldsymbol{\beta}_1$?
```{r}
confint(fit)
```
Response: The 95% confidence interval for β1 1 [-1.034, 2.85]. The true value used to generate the data was β1 = 1.2. Since 1.2 lies within this interval, the confidence interval does capture the true slope. This is expected because a 95% confidence interval should contain the true parameter most of the time, though not in every random sample.

iii. Overlay the LS estimates and the true lines of the mean function onto a copy of the scatterplot you made above.
```{r}
coef(fit)
b0 <- coef(fit)[1]
b1 <- coef(fit)[2]

library(ggplot2)

ggplot(sim_data, aes(x = x, y = y)) +
  geom_point(color = "steelblue", size = 2) +

  # LS fitted regression line
  geom_abline(intercept = b0, slope = b1,
              color = "darkgreen", linewidth = 1) +

  # TRUE regression line
  geom_abline(intercept = 1, slope = 1.2,
              color = "red", linetype = "dashed", linewidth = 1) +

  labs(
    title = "True Line vs Estimated Regression Line",
    subtitle = "Red dashed = true model, Green = LS estimate",
    x = "x",
    y = "y"
  ) +
  theme_minimal()

```

### diagnoses

i. Provide residual plot where fitted $\mathbf{y}$-values are on the x-axis and residuals are on the y-axis. 
```{r}
library(ggplot2)

sim_data$fitted <- fitted(fit)
sim_data$residuals <- resid(fit)

ggplot(sim_data, aes(x = fitted, y = residuals)) +
  geom_point(color = "steelblue", size = 2) +
  geom_hline(yintercept = 0, color = "red", linewidth = 1) +
  labs(
    title = "Residuals vs Fitted Values",
    x = "Fitted values",
    y = "Residuals"
  ) +
  theme_minimal()
```


ii. Provide a normal QQ plot of the residuals.
```{r}
library(ggplot2)

ggplot(sim_data, aes(sample = residuals)) +
  stat_qq(color = "steelblue") +
  stat_qq_line(color = "red", linewidth = 1) +
  labs(
    title = "Normal Q-Q Plot of Residuals"
  ) +
  theme_minimal()
```


iii. Comment on how well the model assumptions are met for the sample you used. 
Response: The residuals versus fitted plot shows the residuals scattered randomly around zero with no clear pattern or curvature, suggesting that the linear relationship between x and y is appropriate. The spread of the residuals appears roughly constant across the range of fitted values, indicating no strong evidence of heteroskedasticity. The normal Q–Q plot shows the points lying close to the reference line, with only small deviations at the extremes, which is consistent with approximately normally distributed errors. Overall, the main assumptions of the linear regression model (linearity, constant variance, and normality of errors) appear to be reasonably well satisfied for this simulated sample.

## Understand sampling distribution and confidence intervals

This part aims to help you understand the notion of sampling statistics and confidence intervals. Let's concentrate on estimating the slope only.  

Generate 100 samples of size $n = 40$, and estimate the slope coefficient from each sample. Also construct 95\% confidence intervals for the slope. 

i. Summarize the LS estimates of the slope. Does the sampling distribution agree with theory? (First specify the theoretical sampling distribution of the LS estimate.)

```{r}
set.seed(1)

B <- 100
n <- 40

slope_est <- numeric(B)
lower <- numeric(B)
upper <- numeric(B)

for(i in 1:B){

  # generate data
  x <- rnorm(n)
  eps <- rnorm(n, 0, 2)
  y <- 1 + 1.2*x + eps

  # fit regression
  fit <- lm(y ~ x)

  # store slope estimate
  slope_est[i] <- coef(fit)[2]

  # store 95% CI
  CI <- confint(fit, "x", level = 0.95)
  lower[i] <- CI[1]
  upper[i] <- CI[2]
}

mean(slope_est)
sd(slope_est)
hist(slope_est, breaks = 15, main = "Sampling Distribution of Slope", xlab = "Slope estimates")
abline(v = 1.2, col = "red", lwd = 2)
```
Response: The theoretical sampling distribution of the least squares slope estimator β̂₁ is normal with mean equal to the true slope β₁ and variance σ² divided by the sum of squared deviations of x from its mean. In symbols, β̂₁ is approximately Normal(β₁, σ² / Σ(xᵢ − x̄)²).

From the simulation, the average of the estimated slopes is about 1.29 and the standard deviation is about 0.297. The histogram of the slope estimates is bell-shaped and centered close to the true value β₁ = 1.2. Although the mean estimate is not exactly 1.2 due to sampling variability from only 100 samples, it is reasonably close. Therefore, the observed sampling distribution agrees well with the theoretical result that the least squares slope estimator is approximately normally distributed around the true slope.

ii.  How many of your 95% confidence intervals capture the true $\boldsymbol{\beta}_1$? Display your confidence intervals graphically. 

```{r}
capture <- (lower <= 1.2) & (upper >= 1.2)

sum(capture)
mean(capture)

plot(1:B, slope_est,
     ylim = range(lower, upper),
     xlab = "Sample number",
     ylab = "Slope estimate and 95% CI",
     main = "95% Confidence Intervals for the Slope")

# draw intervals
for(i in 1:B){
  col_use <- ifelse(capture[i], "black", "red")
  segments(i, lower[i], i, upper[i], col = col_use, lwd = 2)
  points(i, slope_est[i], pch = 19, col = col_use)
}

# true slope
abline(h = 1.2, col = "blue", lwd = 2)
```
Response: Out of the 100 constructed 95% confidence intervals, 95 intervals contain the true slope β₁ = 1.2, giving a coverage proportion of 0.95. This matches the interpretation of a 95% confidence interval: in repeated sampling, about 95% of the intervals will include the true parameter value, while about 5% will miss it due to random sampling variation. The graphical display confirms this, as most intervals cross the horizontal line at 1.2, with only a few (shown in red) failing to cover the true slope.


