---
title: "Modern Data Mining, HW 2"
author:
- Group Member 1
- Group Member 2
- Group Member 3
date: 'Due: 11:59 PM,  Sunday, 02/22'
output:
  html_document:
    code_folding: show
    highlight: haddock
    number_sections: yes
    theme: lumen
    toc: yes
    toc_depth: 4
    toc_float: yes
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: '4'
  word_document:
    toc: yes
    toc_depth: '4'
urlcolor: blue
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width=8, fig.height=4)
options(scipen = 0, digits = 3)  # controls base R output
# check if you have ISLR package, if not, install it
if(!require('pacman')) {install.packages('pacman')}
pacman::p_load(ISLR, tidyverse, data.table) # add the packages needed!!!
```


\pagebreak

# Overview {-}

Principle Component Analysis is widely used in data exploration, dimension reduction, data visualization. The aim is to transform original data into uncorrelated linear combinations of the original data while keeping the information contained in the data. High dimensional data tends to show clusters in lower dimensional view. 

Clustering Analysis is another form of EDA. Here we are hoping to group data points which are close to each other within the groups and far away between different groups. Clustering using PC's can be effective. Clustering analysis can be very subjective in the way we need to summarize the properties within each group. 

Both PCA and Clustering Analysis are so called unsupervised learning. There is no response variables involved in the process. 

For supervised learning, we try to find out how does a set of predictors relate to some response variable of the interest. Multiple regression is still by far, one of the most popular methods. We use a linear model as a working model for its simplicity and interpretability. It is important that we use domain knowledge as much as we can to determine the form of the response as well as the function format of the factors on the other hand. 



## Objectives

- PCA
- SVD
- Clustering Analysis
- Linear Regression

## Review materials

- Study Module 2: PCA
- Study Module 3: Clustering Analysis
- Study Module 4: Multiple regression (Including Simple regression as well)

## Data needed

- `NLSY79.csv`
- `brca_subtype.csv`
- `brca_x_patient.csv`


# Case study 1: Self-esteem 

Self-esteem generally describes a person's overall sense of self-worthiness and personal value. It can play significant role in one's motivation and success throughout the life. Factors that influence self-esteem can be inner thinking, health condition, age, life experiences etc. We will try to identify possible factors in our data that are related to the level of self-esteem. 

In the well-cited National Longitudinal Study of Youth (NLSY79), it follows about 13,000 individuals and numerous individual-year information has been gathered through surveys. The survey data is open to public [here](https://www.nlsinfo.org/investigator/). Among many variables we assembled a subset of variables including personal demographic variables in different years, household environment in 79, ASVAB test Scores in 81 and Self-Esteem scores in 81 and 87 respectively. 

The data is store in `NLSY79.csv`.



Here are the description of variables:

**Personal Demographic Variables**

* Gender: a factor with levels "female" and "male"
* Education05: years of education completed by 2005
* HeightFeet05, HeightInch05: height measurement. For example, a person of 5'10 will be recorded as HeightFeet05=5, HeightInch05=10.
* Weight05: weight in lbs.
* Income87, Income05: total annual income from wages and salary in 2005. 
* Job87 (missing), Job05: job type in 1987 and 2005, including Protective Service Occupations, Food Preparation and Serving Related Occupations, Cleaning and Building Service Occupations, Entertainment Attendants and Related Workers, Funeral Related Occupations, Personal Care and Service Workers, Sales and Related Workers, Office and Administrative Support Workers, Farming, Fishing and Forestry Occupations, Construction Trade and Extraction Workers, Installation, Maintenance and Repairs Workers, Production and Operating Workers, Food Preparation Occupations, Setters, Operators and Tenders,  Transportation and Material Moving Workers
 
 
**Household Environment**
 
* Imagazine: a variable taking on the value 1 if anyone in the respondent’s household regularly read magazines in 1979, otherwise 0
* Inewspaper: a variable taking on the value 1 if anyone in the respondent’s household regularly read newspapers in 1979, otherwise 0
* Ilibrary: a variable taking on the value 1 if anyone in the respondent’s household had a library card in 1979, otherwise 0
* MotherEd: mother’s years of education
* FatherEd: father’s years of education
* FamilyIncome78

**Variables Related to ASVAB test Scores in 1981**

Test | Description
--------- | ------------------------------------------------------
AFQT | percentile score on the AFQT intelligence test in 1981 
Coding | score on the Coding Speed test in 1981
Auto | score on the Automotive and Shop test in 1981
Mechanic | score on the Mechanic test in 1981
Elec | score on the Electronics Information test in 1981
Science | score on the General Science test in 1981
Math | score on the Math test in 1981
Arith | score on the Arithmetic Reasoning test in 1981
Word | score on the Word Knowledge Test in 1981
Parag | score on the Paragraph Comprehension test in 1981
Numer | score on the Numerical Operations test in 1981

**Self-Esteem test 81 and 87**

We have two sets of self-esteem test, one in 1981 and the other in 1987. Each set has same 10 questions. 
They are labeled as `Esteem81` and `Esteem87` respectively followed by the question number.
For example, `Esteem81_1` is Esteem question 1 in 81.

The following 10 questions are answered as 1: strongly agree, 2: agree, 3: disagree, 4: strongly disagree

* Esteem 1: “I am a person of worth”
* Esteem 2: “I have a number of good qualities”
* Esteem 3: “I am inclined to feel like a failure”
* Esteem 4: “I do things as well as others”
* Esteem 5: “I do not have much to be proud of”
* Esteem 6: “I take a positive attitude towards myself and others”
* Esteem 7: “I am satisfied with myself”
* Esteem 8: “I wish I could have more respect for myself”
* Esteem 9: “I feel useless at times”
* Esteem 10: “I think I am no good at all”

## Data preparation

Load the data. Do a quick EDA to get familiar with the data set. Pay attention to the unit of each variable. Are there any missing values? 

```{r quick skim of the data}
 temp <- read.csv('data/NLSY79.csv', header = T, stringsAsFactors = F)
# # missing values? real variables vs. factors? are variable values reasonable?
 str(temp)
 summary(temp)
 levels(as.factor(temp$Job05))
 table(as.factor(temp$Job05))
```


## Self esteem evaluation

**Let concentrate on Esteem scores evaluated in 87. **

0. First do a quick summary over all the `Esteem` variables. Pay attention to missing values, any peculiar numbers etc. How do you fix problems discovered if there is any? Briefly describe what you have done for the data preparation. 

```{r}
esteem87 <- temp[, grep("^Esteem87_", names(temp))]
summary(esteem87)
colSums(is.na(esteem87))
```
Response: We examined the summary statistics of the ten Esteem87 variables and confirmed that all responses fall within the valid range of 1 to 4. There are no missing values in any of the items. The distributions appear reasonable, with positively worded items having lower means and negatively worded items having higher means, consistent with generally moderate to high self-esteem. This will be resolved in the next question.

1. Please note that higher scores on Esteem questions 1, 2, 4, 6, and 7 indicate higher self-esteem, whereas higher scores on the remaining questions suggest lower self-esteem. To maintain consistency, consider reversing the scores of certain Esteem questions. For example, if the esteem data is stored in `data.esteem`, you can use the code `data.esteem[, c(1, 2, 4, 6, 7)] <- 5 - data.esteem[, c(1, 2, 4, 6, 7)]` to invert the scores.

```{r}
esteem87[, c(1,2,4,6,7)] <- 5 - esteem87[, c(1,2,4,6,7)]
```


2. Write a brief summary with necessary plots about the 10 esteem measurements.

```{r}
summary(esteem87)
```
```{r}
boxplot(esteem87, las=2, main="Distribution of Esteem87 Scores")
```

3. Do esteem scores all positively correlated? Report the pairwise correlation table and write a brief summary.

```{r}
cor_mat <- cor(esteem87)
round(cor_mat, 2)
```
Response: The pairwise correlation matrix shows that all ten esteem items are positively correlated after reverse coding. Most correlations are moderate in magnitude, indicating that the items are related but not identical. This suggests that the questions measure a common underlying construct of self-esteem while still capturing slightly different aspects of it.

4. PCA on 10 esteem measurements. (centered but no scaling)

    a) Report the PC1 and PC2 loadings. Are they unit vectors? Are they orthogonal? 
  
```{r pca}
pca <- prcomp(esteem87, center=TRUE, scale.=FALSE)
summary(pca)
```
```{r pc1 and pc2 loading}
pca$rotation[,1:2]
```
```{r unit vector check}
sqrt(sum(pca$rotation[,1]^2))
sqrt(sum(pca$rotation[,2]^2))
```
Response: both PC1 and PC2 have the length (l-2 norm) of 1, suggesting both are unit vectors.
  
```{r orthogonal check}
sum(pca$rotation[,1] * pca$rotation[,2])
```
Response: the dot product of PC1 and PC2 is approximately zero, suggesting two vectors are orthogonal
  
    b) Are there good interpretations for PC1 and PC2? (If loadings are all negative, take the positive loadings for the ease of interpretation)
Response: PC1 has positive and similar loadings across all ten items, suggesting it represents overall self-esteem. Individuals with higher PC1 scores tend to report consistently higher self-esteem across all questions. PC2 contrasts items 1–7 with items 8–10, indicating a secondary dimension that distinguishes general self-confidence from stronger feelings of insecurity or self-doubt. Thus, PC1 captures overall level, while PC2 reflects differences in the type of self-evaluative responses.
  
    c) How is the PC1 score obtained for each subject? Write down the formula.

Reponse: $PC1_i = a_1*x_i1 + a_2*x_i2 + ... + a_{10}*x_i10$
Here, $a_1, ..., a_{10}$ equals PC1 loadings

    d) Are PC1 scores and PC2 scores in the data uncorrelated? 
```{r pc1 and pc2 correlation}
cor(pca$x[,1], pca$x[,2])
```
Response: As expected from PCA theory, PC1 and PC2 scores are uncorrelated.

    e) Plot PVE (Proportion of Variance Explained) and summarize the plot. 

```{r pve}
pve <- pca$sdev^2 / sum(pca$sdev^2)

plot(pve, type="b", xlab="Principal Component", ylab="PVE")
```
Response" The PVE plot shows a sharp drop from PC1 to PC2, with PC1 alone explaining about 46.6% of the total variance. After PC2, the proportion of variance explained decreases gradually, and each remaining component contributes relatively little. This suggests that the data are largely driven by one dominant underlying factor, with a smaller secondary dimension.

    f) Also plot CPVE (Cumulative Proportion of Variance Explained). What proportion of the variance in the data is explained by the first two principal components?

```{r}
plot(cumsum(pve), type="b",
     xlab="Principal Component",
     ylab="Cumulative PVE")
```
Response: The first two principal components explain approximately 60% of total variance, suggesting a strong explanability.

    g) PC’s provide us with a low dimensional view of the self-esteem scores. Use a biplot with the first two PC's to display the data.  Give an interpretation of PC1 and PC2 from the plot. 

```{r}
biplot(pca, scale = 0, xlabs = rep(".", nrow(esteem87)))
```
Response: The biplot shows that all ten items load positively and in a similar direction along PC1, indicating that PC1 represents overall self-esteem. Individuals located further to the right on PC1 tend to report consistently higher self-esteem across all questions. PC2 separates items 1–7 from items 8–10, suggesting that it captures a secondary contrast between general self-confidence and stronger feelings of insecurity. Thus, PC1 reflects overall level of self-esteem, while PC2 distinguishes different patterns in self-evaluative responses.

5. Apply k-means to cluster subjects on the original esteem scores

    a) Find a reasonable number of clusters using within sum of squared with elbow rules.

```{r}
wss <- numeric(10)

for(k in 1:10){
  wss[k] <- sum(kmeans(esteem87, centers=k, nstart=20)$withinss)
}

plot(1:10, wss, type="b")
```
Response: The WSS plot shows a sharp decrease from k = 1 to k = 3, after which the reduction in within-cluster sum of squares becomes more gradual. This suggests that 2 or 3 clusters is a reasonable choice according to the elbow rule.

    b) Can you summarize common features within each cluster?
    
```{r}
k3 <- kmeans(esteem87, centers=3, nstart=20)
aggregate(esteem87, by=list(cluster=k3$cluster), mean)
```
Response: Cluster 1 represents individuals with consistently high self-esteem across all items. Cluster 2 shows uniformly lower scores, indicating comparatively lower self-esteem. Cluster 3 appears to have generally high self-esteem on most items but relatively weaker scores on items 8–10, suggesting moderate overall self-esteem with more variation in feelings of insecurity.
    
    c) Can you visualize the clusters with somewhat clear boundaries? You may try different pairs of variables and different PC pairs of the esteem scores.

```{r}
plot(pca$x[,1:2], col=k3$cluster, pch=19)
```


6. We now try to find out what factors are related to self-esteem? PC1 of all the Esteem scores is a good variable to summarize one's esteem scores. We take PC1 as our response variable. 

    a) Prepare possible factors/variables:

      - EDA the data set first. 

```{r eda}
summary(temp)
colSums(is.na(temp))
```
```{r remove zero and negative income}
temp <- temp[temp$Income87 > 0, ]
temp$logIncome87 <- log(temp$Income87)
```

```{r remove zero and negative height}
temp <- temp[temp$HeightFeet05 > 0, ]
```

      - Personal information: gender, education (05), log(income) in 87, job type in 87.  One way to summarize one's weight and height is via Body Mass Index which is defined as the body mass divided by the square of the body height, and is universally expressed in units of kg/m². Note, you need to create BMI first. Then may include it as one possible predictor. 

```{r create bmi}
height_m <- (temp$HeightFeet05*12 + temp$HeightInch05) * 0.0254
weight_kg <- temp$Weight05 * 0.453592
temp$BMI <- weight_kg / (height_m^2)
```

```{r convert categorical variables}
temp$Gender <- factor(temp$Gender)
```

      - Household environment: Imagazine, Inewspaper, Ilibrary, MotherEd, FatherEd, FamilyIncome78. Do set indicators `Imagazine`, `Inewspaper` and `Ilibrary` as factors. 

```{r factorize}
temp$Imagazine <- factor(temp$Imagazine)
temp$Inewspaper <- factor(temp$Inewspaper)
temp$Ilibrary <- factor(temp$Ilibrary)
```

      - You may use PC1 of ASVAB as level of intelligence

```{r asvab pc1 intelligence}
asvab_vars <- temp[, c("AFQT","Coding","Auto","Mechanic","Elec",
                       "Science","Math","Arith","Word","Parag","Number")]

pca_asvab <- prcomp(asvab_vars, center=TRUE)
temp$ASVAB_PC1 <- pca_asvab$x[,1]
```


```{r pc1 as response variable}
esteem87 <- temp[, grep("^Esteem87_", names(temp))]
pca <- prcomp(esteem87, center=TRUE, scale.=FALSE)
temp$pc1_esteem <- pca$x[,1]
```

    b)   Run a few regression models between PC1 of all the esteem scores in 87 and suitable variables listed in a). Find a final best model with your **own clearly defined criterion**. 

```{r personal varaibles only}
m1 <- lm(pc1_esteem ~ Gender + Education05 + logIncome87 + BMI,
         data=temp)
summary(m1)
```

```{r household only}
m2 <- lm(pc1_esteem ~ Imagazine + Inewspaper + Ilibrary +
         MotherEd + FatherEd + FamilyIncome78,
         data=temp)
summary(m2)
```

```{r add intelligence}
m3 <- lm(pc1_esteem ~ Gender + Education05 + logIncome87 + BMI +
         Imagazine + Inewspaper + Ilibrary +
         MotherEd + FatherEd + FamilyIncome78 +
         ASVAB_PC1,
         data=temp)
summary(m3)
```

```{r}
AIC(m1, m2, m3)
```

Response: To determine the final model, we first fitted several regression models including personal variables, household environment variables, and intelligence (ASVAB_PC1). we compared models using Adjusted R² and AIC, and adopted the principle of parsimony as my selection criterion. Specifically, we selected the model with the lowest AIC while retaining statistically significant predictors.The results indicate that higher education, higher income, and higher intelligence are all positively associated with self-esteem.

      - How did you land this model? Run a model diagnosis to see if the linear model assumptions are reasonably met. 
      
```{r}
par(mfrow=c(2,2))
plot(m3)
```
Response: To assess whether the linear regression assumptions are satisfied, we examined the standard diagnostic plots for the final model. The residuals vs fitted plot does not show strong nonlinearity, suggesting the linearity assumption is reasonable. The Q–Q plot indicates some deviation from normality in the tails, though this is not severe given the large sample size. The scale-location plot suggests mild heteroscedasticity, but the variance pattern is not extreme. The leverage plot shows no highly influential observations. Overall, the linear regression assumptions are reasonably met, with only minor departures from normality and constant variance.

      - Write a summary of your findings. In particular, explain what and how the variables in the model affect one's self-esteem. 
        
Response:Using PC1 of the 1987 self-esteem measures as the response variable, we fitted several regression models and selected a parsimonious final model including Education05, logIncome87, and ASVAB_PC1 based on AIC and statistical significance.
  The results show that all three variables are positively and significantly associated with self-esteem. Specifically, (1) Education05 has a positive coefficient, indicating that individuals with more years of education tend to report higher self-esteem. (2) logIncome87 also has a strong positive effect, suggesting that higher income is associated with higher self-esteem. (3) ASVAB_PC1, representing intelligence, is positively related to self-esteem, meaning individuals with higher cognitive ability tend to have higher self-esteem.
  Among these predictors, income appears to have the strongest practical effect, followed by education and intelligence. The model explains about 10% of the variation in self-esteem, which is reasonable given that self-esteem is influenced by many psychological and social factors not captured in the dataset.
Overall, the findings suggest that socioeconomic status (education and income) and cognitive ability are important factors associated with higher self-esteem.

# Case study 2: Breast cancer sub-type


[The Cancer Genome Atlas (TCGA)](https://www.cancer.gov/about-nci/organization/ccg/research/structural-genomics/tcga), a landmark cancer genomics program by National Cancer Institute (NCI), molecularly characterized over 20,000 primary cancer and matched normal samples spanning 33 cancer types. The genome data is open to public from the [Genomic Data Commons Data Portal (GDC)](https://portal.gdc.cancer.gov/).
 
In this study, we focus on 4 sub-types of breast cancer (BRCA): basal-like (basal), Luminal A-like (lumA), Luminal B-like (lumB), HER2-enriched. The sub-type is based on PAM50, a clinical-grade luminal-basal classifier. (We had hoped to download the data for control groups for each type of the cancer. But failed to do so. Please let us know if you find the appropriate data.)

* Luminal A cancers are low-grade, tend to grow slowly and have the best prognosis.
* Luminal B cancers generally grow slightly faster than luminal A cancers and their prognosis is slightly worse.
* HER2-enriched cancers tend to grow faster than luminal cancers and can have a worse prognosis, but they are often successfully treated with targeted therapies aimed at the HER2 protein. 
* Basal-like breast cancers or triple negative breast cancers do not have the three receptors that the other sub-types have so have fewer treatment options.

We will try to use mRNA expression data alone without the labels to classify 4 sub-types. Classification without labels or prediction without outcomes is called unsupervised learning. We will use K-means and spectrum clustering to cluster the mRNA data and see whether the sub-type can be separated through mRNA data.

We first read the data using `data.table::fread()` which is a faster way to read in big data than `read.csv()`. 

```{r}
brca <- fread("data/brca_subtype.csv")

# get the sub-type information
brca_subtype <- brca$BRCA_Subtype_PAM50
brca <- brca[,-1]
```

1. Summary and transformation

    a) How many patients are there in each sub-type? 

    b) Randomly pick 5 genes and plot the histogram by each sub-type. 

    c) Clean and transform the mRNA sequences by first remove gene with zero count and no variability and then apply logarithmic transform. 


2. Apply kmeans on the transformed dataset with 4 centers (4 clusters) and output the discrepancy table between the real sub-type `brca_subtype` and the cluster labels.

3. Spectrum clustering: to scale or not to scale?

    a) Apply PCA on the centered and scaled dataset. How many PCs should we use and why? You are encouraged to use `irlba::irlba()`. **In order to do so please review the section about SVD in PCA module.** 
    
    b) Plot PC1 vs PC2 of the centered and scaled data and PC1 vs PC2 of the centered but unscaled data side by side. Should we scale or not scale for clustering process? Why? 

4. Spectrum clustering: center but do not scale the data

    a) Use the first 4 PCs of the centered and unscaled data and apply kmeans. Find a reasonable number of clusters using within sum of squared with the elbow rule.
    
    b) Choose an optimal cluster number and apply kmeans. Compare the real sub-type and the clustering label as follows: Plot scatter plot of PC1 vs PC2. Use point color to indicate the true cancer type and point shape to indicate the clustering label. Plot the kmeans centroids with black dots. Summarize how good is clustering results compared to the real sub-type.
    
    c) Compare the clustering result from applying kmeans to the original data and the clustering result from applying kmeans to 4 PCs. Does PCA help in kmeans clustering? What might be the reasons if PCA helps?
    
    d) Now we have an x patient with breast cancer but with unknown sub-type. We have this patient's mRNA sequencing data. Project this x patient to the space of PC1 and PC2. (Hint: remember we remove some gene with no counts or no variablity, take log and centered, then find its PC1 to PC4 scores) Plot this patient in the plot in b) with a black dot as well. Calculate the Euclidean distance between this patient and each of the centroid of the cluster. (Don't forget the clusters are obtained by using 4 PC's) Can you tell which sub-type this patient might have? 
    
```{r}
x_patient <- fread("data/brca_x_patient.csv")
```


# Case Study: Fuel Efficiency in Automobiles

What determines how fuel efficient a car is? Are Japanese cars more fuel efficient?  To answer thes questions we will build various linear models  using 
 the `Auto` dataset from the book `ISLR`. The original dataset contains information for about 400 different cars built in various years.  To get the data, first install the package ISLR which has been done in the first R-chunk.  The `Auto` dataset should be loaded automatically.  Original data source is here: https://archive.ics.uci.edu/ml/datasets/auto+mpg

Get familiar with this dataset first. Tip: you can use the command `?ISLR::Auto` to view a description of the dataset. Our response variable will me `MPG`: miles per gallon.

## EDA


```{r}
# auto <- ISLR::Auto
# ?ISLR::Auto
# summary(auto) 
# hist(auto$acceleration) # to check if there are some outliers
```
a) Explore the data, list the variables with clear definitions. Set each variable with its appropriate class. For example `origin` should be set as a factor. 

b) How many cars are included in this data set? 

c) EDA, focus on pairwise plots and summary statistics. Briefly summarize your findings and any peculiarities in the data.


## What effect does `time` have on `MPG`?

a) Start with a simple regression of `mpg` vs. `year` and report R's `summary` output. Is `year` a significant variable at the .05 level? State what effect `year` has on `mpg`, if any, according to this model. 

b) Add `horsepower` on top of the variable `year` to your linear model. Is `year` still a significant variable at the .05 level? Give a precise interpretation of the `year`'s effect found here. 

c) The two 95% CI's for the coefficient of year differ among (a) and (b). How would you explain the difference to a non-statistician?

d) Create a model with interaction by fitting `lm(mpg ~ year * horsepower)`. Is the interaction effect significant at .05 level? Explain the year effect (if any). 

## Categorical predictors

Remember that the same variable can play different roles! Take a quick look at the variable `cylinders`, and try to use this variable in the following analyses wisely. We all agree that a larger number of cylinders will lower mpg. However, we can interpret `cylinders` as either a continuous (numeric) variable or a categorical variable.

a) Fit a model that treats `cylinders` as a continuous/numeric variable. Is `cylinders` significant at the 0.01 level? What effect does `cylinders` play in this model?

b) Fit a model that treats `cylinders` as a categorical/factor. Is `cylinders` significant at the .01 level? What is the effect of `cylinders` in this model? Describe the `cylinders` effect over `mpg`. 

c) What are the fundamental differences between treating `cylinders` as a continuous and categorical variable in your models? 

d) Can you test the null hypothesis: fit0: `mpg` is linear in `cylinders` vs. fit1: `mpg` relates to `cylinders` as a categorical variable at .01 level?  


## Results

Final modeling question: we want to explore the effects of each feature as best as possible. You may explore interactions, feature transformations, higher order terms, or other strategies within reason. The model(s) should be as parsimonious (simple) as possible unless the gain in accuracy is significant from your point of view.
  
a) Describe the final model. Include diagnostic plots with particular focus on the model residuals and diagnoses.

b) Summarize the effects found.

c) Predict the `mpg` of the following car: A red car built in the US in 1983 that is 180 inches long, has eight cylinders, displaces 350 cu. inches, weighs 4000 pounds, and has a horsepower of 260. Also give a 95% CI for your prediction.


# Simple Regression through simulations 
    
## Linear model through simulations

This exercise is designed to help you understand the linear model using simulations. In this exercise, we will generate $(x_i, y_i)$ pairs so that all linear model assumptions are met.

Presume that $\mathbf{x}$ and $\mathbf{y}$ are linearly related with a normal error $\boldsymbol{\varepsilon}$ , such that $\mathbf{y} = 1 + 1.2\mathbf{x} + \boldsymbol{\varepsilon}$. The standard deviation of the error $\varepsilon_i$ is $\sigma = 2$. 

We can create a sample input vector ($n = 40$) for $\mathbf{x}$ with the following code:

```{r, eval = F, echo = TRUE}
# Generates a vector of size 40 with equally spaced values between 0 and 1, inclusive
x <- seq(0, 1, length = 40)
```


### Generate data

Create a corresponding output vector for $\mathbf{y}$ according to the equation given above. Use `set.seed(1)`. Then, create a scatterplot with $(x_i, y_i)$ pairs. Base R plotting is acceptable, but if you can, please attempt to use `ggplot2` to create the plot. Make sure to have clear labels and sensible titles on your plots.


### Understand the model
i. Find the LS estimates of $\boldsymbol{\beta}_0$ and $\boldsymbol{\beta}_1$, using the `lm()` function. What are the true values of $\boldsymbol{\beta}_0$ and $\boldsymbol{\beta}_1$? Do the estimates look to be good? 

ii. What is your RSE for this linear model fit? Is it close to $\sigma = 2$? 

ii. What is the 95% confidence interval for $\boldsymbol{\beta}_1$? Does this confidence interval capture the true $\boldsymbol{\beta}_1$?

iii. Overlay the LS estimates and the true lines of the mean function onto a copy of the scatterplot you made above.


### diagnoses

i. Provide residual plot where fitted $\mathbf{y}$-values are on the x-axis and residuals are on the y-axis. 

ii. Provide a normal QQ plot of the residuals.

iii. Comment on how well the model assumptions are met for the sample you used. 



## Understand sampling distribution and confidence intervals

This part aims to help you understand the notion of sampling statistics and confidence intervals. Let's concentrate on estimating the slope only.  

Generate 100 samples of size $n = 40$, and estimate the slope coefficient from each sample. Also construct 95\% confidence intervals for the slope. 

i. Summarize the LS estimates of the slope. Does the sampling distribution agree with theory? (First specify the theoretical sampling distribution of the LS estimate.)

ii.  How many of your 95% confidence intervals capture the true $\boldsymbol{\beta}_1$? Display your confidence intervals graphically. 




